---
title: "Project 2"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

**Due:** Monday, April 25, 12:00pm (noon) on Canvas

**Submission:** A template file is provided, with a rough outline: [project_2_template.Rmd](https://sta279-s22.github.io/projects/project_2_template.Rmd)

Once you have completed the report, submit *both* your 
knitted HTML file and the R Markdown file to Canvas.

This project must be completed independently. See "Collaboration" below for more details.

# Data

Biologists hypothesize that birds which live over a small area are less genetically diverse than birds which live over a large area, because a larger habitat allows species to be adapted to a wider range of environments, and because larger habitats help prevent interbreeding. Other features which are believed to be related to genetic diversity include body mass (because body mass is related to population size), and how far from the equator the species lives (because distance from the equator is related to the evolutionary history of the species). Furthermore, researchers are interested in differences between migratory and non-migratory species.

In this project, you will work with data from 387 different bird species, from [this paper](https://www.journals.uchicago.edu/doi/10.1086/705346) exploring bird genetic diversity.

## Variables

The dataset provided for this project contains the following columns:

* `Species`: the scientific name for the bird species. Each row in the dataset represents one species.
* `Family`: the family to which the species belongs. Families are groups of related species.
* `Heterozygosity`: a measure of genetic richness. Birds (and most animals) have two copies of each chromosome; expected heterozygosity is the probability that for a randomly selected gene, these two copies will be different versions (alleles) of the gene.
* `Allelic.richness`: The average number of different alleles (versions of a gene) for each gene. 
* `Breeding.range.size`: Size of the spatial range over which this bird breeds, in units of 10,000 square km.
* `Body.mass`: Average body mass of the species (grams).
* `Latitude`: Mid-point latitude of the species breeding range. (All latitudes are positive, even those in the southern hemisphere, so Latitude measures distance from the equator).
* `Migratory.status`: whether the species is Migratory or Resident (non-migratory).

## Downloading the data

To load the data, use the code below. It will import a data set called `birds` into R.

```r
birds <- read_csv("https://sta279-s22.github.io/projects/birds.csv")
```


# Research questions

The biologists who collected this data are interested in the following questions:

1. Breeding range is likely related to migratory status, since migratory birds travel more. Does the relationship between breeding range and genetic diversity depend on migratory status, after accounting for body mass and latitude? You should account for correlation between species from the same family.
2. Is there systematic variation in genetic diversity between families, after accounting for differences due to breeding range, body mass, latitude, and migratory status?

To address these research questions, you will perform exploratory data analysis, fit one or more models, and analyze the results. You may use heterozygosity as your measure of genetic diversity.

# What will you be turning in?

You will submit two documents on Canvas for this project.

* Your written report (.html file):
  * This is the knitted write up that will explain your work, and the answers to the researchers' questions
  * This report will contain very specific sections, which you must label and include in your final report. More details on the sections needed for this report are included below.
  * In your formal report, there should be **no code** showing. This
includes warnings and other stray code output – hide it all.
  * You will be graded on writing as well as your statistical analysis.
* Your R Markdown (.Rmd) file, with all the code you used for your analysis
  * The goal is that a person who reads your report, and wants to
replicate your results, could look at your .Rmd file and completely reproduce the results and figures in the knitted report.

# Content

Your report should contain the following labeled sections.

## Abstract

This is the first section in your report, but it is actually the last thing you will write. Called an abstract in academia, and an executive summary in industry, this one paragraph summary of your entire paper is the first thing people will read. For example:

*Estimating the number of pandas inhabiting a national panda preserve is critical to understanding the health of the panda species as a whole. In this report, we discuss the process of estimating the number of pandas in the Wolong National Nature Reserve. Data were collected from volunteers walking trails in the park, and we applied capture-recapture estimation process to then estimate the total number of pandas in the park. We detail the steps of the process, and compare our results to two other possible methods of estimating pandas. Based on our estimation, there are between 140 and 155 pandas in the surveyed area. We discuss our findings, and limitations of the study, in the following paper.*
  
## Section 1: Introduction

Write 2-3 paragraphs in which you:

* Motivate the research questions
* Describe where the data came from
* Provide background on the dataset
* Finish by summarizing what you will do in the report

## Section 2: Data

In this section, you will summarize and explore your data. First, write 1-2 paragraphs which:

* Describe details of the data. What does a row in the data represent? How many rows and columns? What information do the variables record? Are there any missing data?
* Describe any data manipulation. Did you remove any missing values? Did you focus only on a subset of the data? Did you create any new variables?

Next, perform exploratory data analysis. Your goal is to create any visualizations necessary to exploring the data before fitting a model (e.g. tables, histograms, scatterplots, etc.). Make sure that:

* You have at least three graphs shown
* Your figures have axis labels, captions, and are numbered Figure 2.1, Figure 2.2, etc., where the 2. represents the section number, and the number after the decimal represents how many plots so far in this section
* Any plot that is shown should be discussed in writing in your report, and referenced by number (e.g., "As shown in Figure 2.1...").  What information does this graph give us, and why is that important for us to know when before we start building our model in the next section?

## Section 3: Modeling

In this section, you will build one or more models to answer the two research questions. For each model, you will:

* Write down a population model that allows you to answer a research question, and explain why the model is appropriate for these data.
* Fit the model, and interpret any relevant parameter estimates or statistics.
* Check assumptions for the model. Make sure to include relevant diagnostic plots, and perform any necessary transformations to address assumption violations.

You will then use your fitted model(s) to perform hypothesis tests that allow you to address the research questions. When testing hypotheses, make sure to:

* State the hypotheses in terms of one or more model parameters
* Calculate a test statistic and p-value
  * For testing fixed effect terms (i.e., $\beta$s ) you should use a nested F statistic. You may calculate a p-value using either an approximate F distribution (Satterthwaite or Kenward-Roger methods) or parametric bootstrapping.
  * For testing random effect terms (i.e., $\sigma_u^2$ ) you should use parametric bootstrapping. The estimated intra-class correlation is an acceptable test statistic.
* Use the p-value to make a conclusion about the research question


## Section 4: Discussion

In this section, you are providing a conclusion. Write 1–2 paragraphs which summarize what you learned from your analysis, and how it addresses the original research questions. Also discuss any limitations to your analysis, and what you other relationships you might explore in future.
  
# Appearance and style

### Writing

The report should be written like an article or research paper: in full sentences and paragraphs, with headings for each section. You should not write your report with question numbers or as a list of bullet points. Scientific articles are generally written in third person, though "we" can also be acceptable ("we can see from Figure 2.1 ...") in some disciplines.

### Code

As with previous assignments, you will write the document in R Markdown. However, in full reports, the only output that should be visible from code chunks are figures and tables. If a code chunk does not produce a figure or table, you can hide it from the knitted document with `include=F`:

````
```{r, include=F}`r ''`

```
````

If a code chunk produces a figure or table, only the figure or table should be visible in the knitted document. You can hide the chunk but display the output with `echo=F, message=F, warning=F`:

````
```{r, echo=F, message=F, warning=F}`r ''`

```
````

### Figures

Figures should have labeled axes, and should be clear and easy to read. Figures should also be captioned and numbered; to caption a figure, use `fig.cap = "..."` in the chunk options. For example (scroll to the right on the code to see it all),

````
```{r, echo=F, message=F, warning=F, fig.cap="Figure 2.1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}`r ''`
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```
````

is displayed as

```{r include=F}
library(tidyverse)
library(palmerpenguins)
```

```{r, echo=F, message=F, warning=F, fig.cap="Figure 1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```

Captions should provide enough information to understand what is being plotted, but interpretation can be left to the main text. Refer to figures by their number in the text. Make sure that any figures you include are discussed in the text.

### Tables

Tables should be nicely formatted, and have a number and caption. This can be done with the `kable` function in the `knitr` package.

For example,

````
```{r, echo=F, message=F}`r''`
table(penguins$island, penguins$species) %>%
  knitr::kable(caption = "Table 3.2: Penguins by island and species")
```
````

is displayed as

```{r, echo=F, message=F}
table(penguins$island, penguins$species) %>%
  knitr::kable(caption = "Table 3.2: Penguins by island and species")
```

### Writing math in R Markdown

If you want to write mathematical notation, we need to tell Markdown, “Hey, we’re going to make a math symbol!” To do that, you use dollar signs. For instance, to make $\widehat{\beta}_1$, you simply put `$\widehathat{\beta}_1$` into the white space (not a chunk) in your Markdown.

Here are some examples of writing math, which you can adapt:

| Math | Code |
| --- | --- |
| $Y_i \sim Bernoulli(\pi_i)$ | `$Y_i \sim Bernoulli(\pi_i)$` |
| $\log \left( \dfrac{\pi_i}{1 - \pi_i} \right)$ | `$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right)$` |
| $\widehat{\pi}_i$ | `$\widehat{\pi}_i$` |

# Collaboration

One purpose of the projects is to capture your knowledge and understanding of statistics and data analysis. Unlike homework and lab assignments, this project must be completed on your own, without collaborating with other students. 

You may:

* ask the instructor for help
* use resources from the class notes and internet (Wikipedia, StackExchange, etc.)

You may not:

* work with another student on the project
* ask a TA or tutor (e.g., at the MSC) for help
