---
title: Logistic regression interpretation and assumptions
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---


### Agenda

.large[
* Today:
  * Recap logistic regression model
  * Interpreting coefficients
  * Model assumptions
* Reminders: 
  * Lab 1 due Friday on Canvas
  * Office hours today and tomorrow
  * Extension policy
* MSC hours:
  * TA: Ethan Lewis
  * Mondays 7 - 9pm, Kirby 103
]

---

### Data

.large[
Data on 400 undergraduate students from different universities, and the outcome of their application to graduate school. Variables include:

* *admit*: whether or not the student was admitted (0 = no, 1 = yes)
* *gre*: the student’s score on the GRE (graduate record exam)
* *gpa*: the student’s grade point average
* *rank*: how prestigious the student’s undergraduate institution is (1 = most prestigious, 2, 3, 4 = least prestigious)

.question[
What is the relationship between a student's GPA and whether they are accepted to graduate school?
]
]

---

### EDA

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

grad_app %>%
  ggplot(aes(x = as.factor(admit), y = gpa)) +
  geom_boxplot() +
  labs(x = "Admitted to graduate school?",
       y = "GPA") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.large[
.question[
Students who were accepted to graduate school tend to have higher GPAs than students who were not accepted.
]
]

---

### Logistic regression model

.large[
* $Y_i =$ whether $i$th student was accepted (0 = no, 1 = yes)
* $\pi_i = P(Y_i = 1)$
* $GPA_i =$ GPA of $i$th student
]

.large[
What is my logistic regression model?

.abox[
(A) $\hspace{0.2cm} Y_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \pi_i = \beta_0 + \beta_1 GPA_i$
]

.bbox[
(B) $\hspace{0.2cm} Y_i \sim Normal(\pi_i) \hspace{0.5cm} \dfrac{\pi_i}{1 - \pi_i} = \beta_0 + \beta_1 GPA_i$
]

.cbox[
(C) $\hspace{0.2cm} Y_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GPA_i$
]
]

--

.large[
**Answer:** (C)
]

---

### Logistic regression model

.large[
$Y_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GPA_i$

* $\pi_i =$ probability of acceptance
* $\dfrac{\pi_i}{1 - \pi_i} =$ odds of acceptance
* $\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) =$ log odds of acceptance

]

---

### Concept check

.large[
The equation of the fitted logistic regression model is

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.36 + 1.05 \ GPA_i$

.question[
As GPA increases, how does the estimated probability of acceptance to graduate school change?
]

.abox[
(A) It increases
]

.bbox[
(B) It decreases
]

.cbox[
(C) It stays the same
]
]

--

.large[
**Answer:** (A) It increases (because the estimated slope, 1.05, is positive)
]

---

### Concept check

.large[

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.36 + 1.05 \ GPA_i$

.question[
What are the estimated **log odds** of acceptance for a student with a GPA of 3.0?
]

.abox[
(A) -4.36
]

.bbox[
(B) 1.05
]

.cbox[
(C) -1.21
]

.dbox[
(D) 0.30
]
]

--

.large[
**Answer:** $-4.36 + 1.05 \cdot 3.0 = -1.21$
]

---

### Concept check

.large[

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.36 + 1.05 \ GPA_i$

.question[
What are the estimated **odds** of acceptance for a student with a GPA of 3.0?
]

.abox[
(A) 0.013
]

.bbox[
(B) 2.86
]

.cbox[
(C) -1.21
]

.dbox[
(D) 0.30
]
]

--

.large[
**Answer:** $\exp\{-1.21\} = 0.30$
]

---

### Concept check

.large[

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.36 + 1.05 \ GPA_i$

.question[
The estimated odds of acceptance for a student with a GPA of 3.0 are 0.30. How would I interpret those odds?
]

.abox[
(A) The probability of acceptance is 30%
]

.bbox[
(B) The probability of being accepted is 0.3 times the probability of being rejected
]

.cbox[
(C) The probability of being accepted with a GPA of 3.0 is 0.3 times the probability of being accepted with a GPA of 4.0
]
]

--

.large[
**Answer:** (B)
]

---

### Concept check

.large[

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.36 + 1.05 \ GPA_i$

.question[
If GPA increases by one unit, what is the associated change in our estimated log odds of acceptance?
]

.abox[
(A) An increase of 1.05
]

.bbox[
(B) A decrease of 1.05
]

.cbox[
(C) An increase of 4.36
]

.dbox[
(D) A decrease of 4.36
]
]

--

.large[
**Answer:** (A)
]

---

### Interpretation

.large[

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 \ X_i$

$\widehat{\beta}_1 =$ the change in predicted log odds associated with a one unit increase in $X$

.question[
Unfortunately, log odds aren't very natural to think about. Can we interpret $\widehat{\beta}_1$ in terms of *odds* instead? 
]
]

---

### Odds ratio

.large[
$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.36 + 1.05 \ GPA_i$

* If $GPA_i = 3.0$, the estimated odds are $\exp\{-4.36 + 1.05 \cdot 3.0\} = 0.298$
* If $GPA_i = 4.0$, the estimated odds are $\exp\{-4.36 + 1.05 \cdot 4.0\} = 0.852$

The **odds ratio** is $\dfrac{0.852}{0.298} = 2.86$

* The odds of acceptance are 2.86 times higher for a student with a GPA of 4.0, compared to a student with a GPA of 3.0
]

---

### Concept check

.large[
$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.36 + 1.05 \ GPA_i$

.question[
What is the odds ratio (of acceptance) for a student with a GPA of 3.5, compared to a student with a GPA of 2.5?
]

.abox[
(A) 0.50
]

.bbox[
(B) 0.18
]

.cbox[
(C) 2.86
]
]

--

.large[
**Answer:** $\dfrac{\exp\{ -4.36 + 1.05 \cdot 3.5\}}{\exp\{ -4.36 + 1.05 \cdot 2.5\}} = 2.86$
]

---

### Interpretation

.large[
Fitted model: log odds form

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -4.36 + 1.05 \ GPA_i$

* For every one unit increase in GPA, we predict that the log odds of acceptance increase by 1.05

Fitted model: odds form

$\dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} = e^{ -4.36 + 1.05 \ GPA_i } = e^{-4.36} e^{1.05 \ GPA_i}$

* For every one unit increase in GPA, we predict that the odds of acceptance get multiplied by $\exp\{1.05\} = 2.86$
]

---

### Interpretation

.large[

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 \ X_i$

* $\widehat{\beta}_1 =$ the (additive) change in predicted log odds associated with a one unit increase in $X$ 
* $\exp\{\widehat{\beta}_1\} =$ the (multiplicative) change in predicted odds associated with a one unit increase in $X$

.question[
What about $\widehat{\beta}_0$ and $\exp\{\widehat{\beta}_0\}$?
]
]

--

.large[
The estimated log odds and odds when $X_i = 0$ (doesn't always make sense)
]

---

### Class Activity

.large[
[https://sta279-s22.github.io/class_activities/ca_lecture4.html](https://sta279-s22.github.io/class_activities/ca_lecture4.html)
]

---

### Class Activity

.large[
$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -2.901 + 0.0036 \ Score_i$

.question[
What is the predicted change in the odds of acceptance associated with a one point increase in GRE score?
]
]

--

.large[
The predicted odds of acceptance are multiplied by $\exp\{0.0036\} = 1.0036$
]

---

### Class Activity

.large[
$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -2.901 + 0.0036 \ Score_i$

.question[
What is the predicted change in the odds of acceptance associated with a 100 point increase in GRE score?
]
]

--

.large[
The predicted odds of acceptance are multiplied by $\exp\{0.0036 \cdot 100\} = 1.433$
]

---

### Class Activity

.large[
.question[
The logistic regression model assumes that the change in odds associated with an increase of 1 point in GRE score is constant. Is the change in *probability* also constant?
]
]

--

.large[
No -- it depends on GRE score.

* Score = 400 $\rightarrow$ probability = 0.19
* Score = 500 $\rightarrow$ probability = 0.25
* Score = 600 $\rightarrow$ probability = 0.32

.pull-left[
Differences:
  * 0.25 - 0.19 = 0.06
  * 0.32 - 0.25 = 0.07
]

.pull-right[
Ratios:
  * 0.25/0.19 = 1.32
  * 0.32/0.25 = 1.28
]
]

---

### Logistic regression assumptions

.large[
$Y_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 X_i$

**Assumption 1 (shape):** The log odds are a linear function of $X_i$
  * As a result, a unit increase in $X$ is associated with a constant (additive) change in the log odds, and a constant (multiplicative) change in the odds

]

---

### Assessing shape: empirical log odds

.large[
Recall the contraceptive survey data:


| Age | 20 | 21 | 22 | 23 | 24 | 25 | ... |
| --- | --- | --- | --- | --- | --- | --- | --- |
| Contraceptive use = 0 | 17 | 19 | 27 | 21 | 24 | 33 | ... |
| Contraceptive use = 1 | 11 | 29 | 32 | 23 | 37 | 47 | ... |
]

---

### Assessing shape: empirical log odds

.large[
Want to know if the log odds of contraceptive use are a linear function of age.

**Step 1:** Calculate empirical log odds of contraceptive use for each age.

| Age | 20 | 21 | 22 | 23 | 24 | 25 | ... |
| --- | --- | --- | --- | --- | --- | --- | --- |
| Contraceptive use = 0 | 17 | 19 | 27 | 21 | 24 | 33 | ... |
| Contraceptive use = 1 | 11 | 29 | 32 | 23 | 37 | 47 | ... |
| Empirical log odds | log(11/17) = -0.435 | 0.423 | 0.170 | 0.091 | 0.433 | 0.354 | ... |
]

---

### Assessing shape: empirical log odds

.large[
Want to know if the log odds of contraceptive use are a linear function of age.

**Step 2:** Plot empirical log odds against age.

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
cmc_data <- read_csv("cmc.data.txt", col_names = F)
colnames(cmc_data) <-c("WifeAge", "WifeEdu", "HusbandEdu", "NumChild", "WifeRel", "WifeWork", "HusbandOcc","Living", "Media", "Choice")
cmc_data$Choice <- ifelse( cmc_data$Choice == 1, "None", ifelse(cmc_data$Choice==2, "Short", "Long"))
cmc_data$ChoiceBin <- ifelse(cmc_data$Choice =="None", 0 , 1)
cmc_data <- cmc_data %>% 
  filter(WifeAge >= 20, WifeAge <= 35)

cmc_data %>%
  group_by(WifeAge) %>%
  summarize(prop = mean(ChoiceBin),
          log_odds = log(prop/(1 - prop))) %>%
  ggplot(aes(x = WifeAge, y = log_odds)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se=F, lwd = 1.5) +
  labs(x = "Age", y = "Empirical log odds") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

The shape assumption is reasonable if the plot looks roughly linear.
]

---

### Empirical log odds plots

.large[
Now let's try this with the grad school admissions data:

| GPA | 2.26 | 2.42 | 2.48 | 2.52 | 2.55 | 2.56 | ... |
| --- | --- | --- | --- | --- | --- | --- | --- |
| Admit = 0 | 1 | 1 | 1 | 1 | 1 | 1 | ... |
| Admit = 1 | 0 | 1 | 0 | 0 | 0 | 0 | ... |

.question[
What problem do I run into?
]
]

--

.large[
There are a small number of observations for each GPA, so we can't calculate empirical log odds for each GPA.

**Solution:** Divide the data into *bins*
]

---

### Empirical log odds plots

.large[
* Divide observations into bins of equal size, based on values of predictor $X$
* Calculate empirical log odds in each bin
* Plot empirical log odds

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
num_bins <- 8

logodds_table <- grad_app %>%
  mutate(obs = admit,
         bin = cut(gpa, 
                   breaks = quantile(gpa, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin) %>%
  summarize(mean_x = mean(gpa),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "GPA",
       y = "Empirical log odds",
       color = "",) +
  theme(text = element_text(size=20))
```
]
