---
title: Poisson Regression 
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Project 2 due April 25
* No assignments due this week
* No class Friday (holiday)
* Remaining course components:
  * Project 2
  * Lab 9
  * Final exam
* Today: start Poisson regression
]

---

### Data

.large[
2015 Family Income and Expenditure Survey (FIES) on households in the Phillipines. Variables include

* `age`: age of the head of household
* `numLT5`: number in the household under 5 years old
* `total`: total number of people other than head of household
* `roof`: type of roof (stronger material can sometimes be used as a proxy for greater wealth)
* `location`: where the house is located (Central Luzon, Davao Region, Ilocos Region, Metro Manila, or Visayas)
]

---

### Data

.large[
Questions:

* How is the age of head of household related to the number of people in the household?
* Is the type of roof material related to the number of people in the household?

.question[
To answer these questions, our response variable is `total` (total number of people other than head). What kind of variable is this?
]
]

--

.large[
A count variable (a numeric variable whose values are non-negative integers 0, 1, 2, ...)
]

---

### Building a parametric model

.large[
**Step 1:** Choose a reasonable distribution for $Y$

* $Y_i = total_i$ is a count variable!
* Unfortunately, we don't know any distributions for count data
* Bernoulli and Categorical distributions are for categorical data
* Normal distributions are for continuous data; count data is discrete

.question[
We need a new distribution!
]
]

---

### Exploring the response

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
fies <- read_csv("~/Documents/Teaching/sta279-s22.github.io/slides/fies.csv")

fies %>%
  ggplot(aes(x = `Total Number of Family members` - 1)) +
  geom_bar() +
  labs(x = "Number of family members") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
* **Right skewed, unimodal** distribution
* We can use a *Poisson* distribution
]

---

### Poisson distribution

.large[
If $Y_i \sim Poisson(\lambda)$, then $Y_i$ takes values $y = 0, 1, 2, ...$ with probabilities

$$P(Y_i = y) = \dfrac{e^{-\lambda} \lambda^y}{y!}$$

.question[
Does this distribution look familiar?
]
]

---

### Poisson distribution

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=8, fig.height=6}
library(gridExtra)

x1 <- rpois(1000, 1)
x2 <- rpois(1000, 2)
x3 <- rpois(1000, 5)
x4 <- rpois(1000, 10)

p1 <- data.frame(Count = x1) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Lambda = 1") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p2 <- data.frame(Count = x2) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Lambda = 2") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p3 <- data.frame(Count = x3) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Lambda = 5") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p4 <- data.frame(Count = x4) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Lambda = 10") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

.large[
.question[
How is $\lambda$ related to the distribution?
]
]

---

### Poisson distribution

.large[
If $Y_i \sim Poisson(\lambda)$, then $Y_i$ takes values $y = 0, 1, 2, ...$ with probabilities

$$P(Y_i = y) = \dfrac{e^{-\lambda} \lambda^y}{y!}$$

* $\lambda$ is the mean of the distribution
* $\lambda$ is also the variance! (the mean and variance are the same)
* Our goal is to estimate $\lambda$, just like our goal was to estimate $\pi$ in logistic regression
]

---

### Estimating $\lambda$ with maximum likelihood

.large[
$$P(Y_i = y) = \dfrac{e^{-\lambda} \lambda^y}{y!}$$

Suppose that $Y_1,...,Y_n \overset{iid}{\sim} Poisson(\lambda)$.

.question[
What is the maximum likelihood estimate $\widehat{\lambda}$?
]
]

--

.large[
You saw on a lab that $\widehat{\lambda} = \frac{1}{n} \sum \limits_{i=1}^n Y_i$
]

---

### Class activity

.large[
[https://sta279-s22.github.io/class_activities/ca_lecture_26.html](https://sta279-s22.github.io/class_activities/ca_lecture_26.html)
]

---

### Class activity

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=8, fig.height=6}
library(gridExtra)

set.seed(3)

x1 <- rpois(1000, 15)
x2 <- rpois(1000, 4)
x3 <- rpois(1000, 2)
x4 <- rpois(1000, 7)

p1 <- data.frame(Count = x1) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Graph A") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p2 <- data.frame(Count = x2) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Graph B") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p3 <- data.frame(Count = x3) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Graph C") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p4 <- data.frame(Count = x4) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Graph D") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

.large[
.question[
What do you think $\lambda$ is for each graph?
]
]

---

### Class activity

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=8, fig.height=6}
library(gridExtra)

p1 <- data.frame(Count = x1) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Graph A") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p2 <- data.frame(Count = x2) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Graph B") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p3 <- data.frame(Count = x3) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Graph C") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

p4 <- data.frame(Count = x4) %>%
  ggplot(aes(x = Count)) +
  geom_bar() +
  labs(title = "Graph D") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  scale_x_continuous(limits=c(0, 25))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

.large[
$A: \lambda = 15 \hspace{0.4cm} B: \lambda = 3 \hspace{0.4cm} C: \lambda = 2 \hspace{0.4cm} D: \lambda = 7$
]

---

### Class activity

.large[
$Y_i = \text{number of dogs adopted from animal shelter}$

$Y_i \sim Poisson(1.5)$

.question[
What is the probability that at most two dogs are adopted?
]
]

--

.large[
$\dfrac{e^{-1.5} \cdot 1.5^0}{0!} + \dfrac{e^{-1.5} \cdot 1.5^1}{1!} + \dfrac{e^{-1.5} \cdot 1.5^2}{2!} = 0.809$
]

---

### Class activity

.large[
.question[
The Poisson distribution is for count data. Why is it ok for $\lambda$ to not be a whole number?
]
]

--

.large[
Because $\lambda$ represents the *average* count, which can be a fraction even if the individual counts are whole numbers.
]

---

### Poisson regression

.large[
$Y_i =$ number of people in household other than head

How is $Y_i$ related to the age of the head of the household?

**Step 1:** Choose a reasonable distribution for $Y$

$$Y_i \sim Poisson(\lambda_i)$$

**Step 2:** Choose a model for any parameters

$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i$$
.question[
Why do you think we use $\log(\lambda_i)$ instead of just $\lambda_i$?
]
]

--

.large[
$\lambda_i$ is non-negative, but $\log(\lambda_i)$ can be either positive or negative
]

---

### Poisson regression

.large[
$Y_i =$ number of people in household other than head

**Step 1:** Choose a reasonable distribution for $Y$

$$Y_i \sim Poisson(\lambda_i)$$

**Step 2:** Choose a model for any parameters

$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i$$

.question[
How is this different than linear regression with a log-transformed response?
]
]

---

### Comparing with transformation

.large[
**Poisson regression:**

$$Y_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i$$
**Linear regression with transformed response:**

$$\log(Y_i) \sim N(\mu_i, \sigma^2)$$

$$\mu_i = \beta_0 + \beta_1 Age_i$$

**Difference:** If $\log(Y_i)$ is normal, this does **not** mean that $Y_i$ is Poisson. So we're using different distributions!
]

---

### Comparing with transformation

.large[
**Poisson regression:**

$$Y_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i$$
**Linear regression with transformed response:**

$$\log(Y_i) \sim N(\mu_i, \sigma^2)$$

$$\mu_i = \beta_0 + \beta_1 Age_i$$

**Difference:**

* $\lambda_i =$ average $Y_i$
* $\mu_i =$ average $\log(Y_i)$
* $\log(\text{average } Y_i) \neq \text{average} \log(Y_i)$ (**Jensen's inequality**)
]

---

### Fitting the model

```{r include=F}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
fies <- fies %>%
  mutate(total = `Total Number of Family members`,
         age = `Household Head Age`)
```

.large[
```{r, output.lines=10:12}
m1 <- glm(total ~ age, family = poisson,
          data = fies)
summary(m1)
```

$$\log(\widehat{\lambda}_i) = 1.714 - 0.0035 \ Age_i$$

.question[
How can we interpret the slope?
]
]

---

### Fitting the model

.large[
$$\log(\widehat{\lambda}_i) = 1.714 - 0.0035 \ Age_i$$

For every additional year in age of the head of house, we expect the log of the *average* household size to decrease by 0.0035.

.question[
Can I interpret on the un-logged scale?
]
]

--

.large[
* $\widehat{\lambda}_i = e^{1.714}e^{-0.0035 Age_i}$
* $e^{-0.0035} = 0.9965$
* For every additional year in age of the head of house, we expect the average household size to decrease by 0.35%

]

---

### Assumptions

.large[
$$Y_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i$$

.question[
What assumptions does our Poisson regression model make?
]
]

--

.large[
* **Shape:** The shape of the regression model is correct 
* **Independence:** The observations are independent
* **Poisson distribution:** A Poisson distribution is a good choice for $Y_i$
]

---

### The shape assumption

.large[
**Shape assumption:** The shape of the regression model is correct

.question[
How can I assess this assumption?
]
]

--

.large[
Log empirical mean plot:

* Create bins for explanatory variable
* In each bin, find the log average count of household size
* Plot log average count of household size against age
* Check to see what the shape looks like!
]

---

### Checking the shape assumption

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=7, fig.height=5}
set.seed(6)
fies %>%
  group_by(age) %>%
  summarize(emp_mean = log(mean(total)),
            age_mean = mean(age)) %>%
  ungroup() %>%
  ggplot(aes(x = age_mean, y = emp_mean)) +
  geom_jitter(size = 2, width=0.3, height=0.3) +
  labs(x = "Age", y = "log(Average number of people)") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
What shape seems appropriate?
]
]

---

### Second order polynomial

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=7, fig.height=5}
set.seed(6)
fies %>%
  group_by(age) %>%
  summarize(emp_mean = log(mean(total)),
            age_mean = mean(age)) %>%
  ungroup() %>%
  ggplot(aes(x = age_mean, y = emp_mean)) +
  geom_jitter(size = 2, width=0.3, height=0.3) +
  labs(x = "Age", y = "log(Average number of people)") +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ poly(x,2), lwd=1.2)
```

---

### Poisson distribution assumption

.large[
$$Y_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i + \beta_2 Age_i^2$$

**Poisson distribution assumption:** The Poisson distribution is a good choice for $Y_i$.

.question[
What are some characteristics of the Poisson distribution we could check?
]
]

--

.large[
* $Y_i$ is a count variable
* For a given age, $Y_i$ is unimodal and right skewed
* The mean of $Y_i$ is approximately equal to its variance
]

---

### Checking distribution shape

.large[
Look at the distribution of $Y_i$ for different ages:

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=8, fig.height=6}
fies %>%
  mutate(age_binned = cut(age, breaks = 9)) %>%
  ggplot(aes(x = total)) +
  geom_bar() +
  facet_wrap(~age_binned) +
  theme_bw()
```
]

---

### Checking mean and variance

.large[
```{r}
mean(fies$total)
var(fies$total)
```

* Rule of thumb: not too concerned if variance is less than twice the sample mean
]
