---
title: More multinomial regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Motivation

.large[
**Question:** What is the relationship between age and contraceptive use for women in Indonesia?

**Data:** 1473 Indonesian couples, with variables
* $Y_i$ = contraceptive method used (1 = no use, 2 = long-term, 3 = short-term)
* $X_i$ = Wife's age (numeric)
]

---

### EDA: log relative risk

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=5, fig.align='center'}
library(tidyverse)
library(nnet)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

cmc_data <- read_csv("cmc.data.txt", col_names = F)
colnames(cmc_data) <-c("WifeAge", "WifeEdu", "HusbandEdu", "NumChild", "WifeRel", "WifeWork", "HusbandOcc","Living", "Media", "Choice")
cmc_data$Choice <- ifelse( cmc_data$Choice == 1, "None", ifelse(cmc_data$Choice==2, "Short", "Long"))
cmc_data <- cmc_data %>%
  mutate(Choice = factor(Choice, levels=c("None", "Short", "Long")))
cmc_data %>%
  group_by(WifeAge) %>%
  summarize(count_long = sum(Choice == "Long"),
            count_none = sum(Choice == "None")) %>%
  mutate(log_rr = log(count_long/count_none)) %>%
  ggplot(aes(x = WifeAge, y = log_rr)) +
  geom_point(size=2) +
  labs(x = "Wife's Age", y = "Log RR: Long vs. None") +
  theme_bw() +
  theme(text = element_text(size=25))
```

---

### EDA: log relative risk

```{r echo=F, fig.width=7, fig.height=5, fig.align='center'}
cmc_data %>%
  group_by(WifeAge) %>%
  summarize(count_long = sum(Choice == "Short"),
            count_none = sum(Choice == "None")) %>%
  mutate(log_rr = log(count_long/count_none)) %>%
  ggplot(aes(x = WifeAge, y = log_rr)) +
  geom_point(size=2) +
  labs(x = "Wife's Age", y = "Log RR: Short vs. None") +
  theme_bw() +
  theme(text = element_text(size=25))
```

---

### Multinomial regression model

.large[
**Step 1: Choose a reasonable distribution for $Y$**

$$Y_i \sim Categorical(\pi_{i(None)}, \pi_{i(Short)}, \pi_{i(Long)})$$
**Step 2: Choose a model for any parameters**

$$\log \left( \dfrac{\pi_{i(Long)}}{\pi_{i(None)}} \right) = \beta_{0(Long)} + \beta_{1(Long)} Age_i + \beta_{2(Long)} Age_i^2$$
$$\log \left( \dfrac{\pi_{i(Short)}}{\pi_{i(None)}} \right) = \beta_{0(Short)} + \beta_{1(Short)} Age_i + \beta_{2(Short)} Age_i^2$$
]

---

### Fitting the model in R

.large[
* To fit a multinomial regression, we use the `nnet` package

```r
library(nnet)
```

* Syntax is very similar to other regression techniques

```r
m1 <- multinom(Choice ~ WifeAge + I(WifeAge^2), 
               data = cmc_data)
```
]

---

### Fitting the model in R

.large[
When you fit the model, you get the following output:

```{r}
m1 <- multinom(Choice ~ WifeAge + I(WifeAge^2), 
               data = cmc_data)
```

* This reflects the process used to estimate the model parameters -- we won't get into that here
]

---

### Fitting the model in R

.large[
```{r}
summary(m1)
```
]

---

### Fitting the model in R

.large[
```{r, echo=F, output.lines=4:7}
summary(m1)
```
]

.large[
.question[
What is the fitted model for short term vs. no contraceptive use?
]
]

--

.large[

$$\log \left( \dfrac{\widehat{\pi}_{i(Short)}}{\widehat{\pi}_{i(None)}} \right) = -8.234 + 0.456 Age_i - 0.0065 Age_i^2$$
]

---

### Class activity, Part I

.large[
[https://sta279-s22.github.io/class_activities/ca_lecture_15.html](https://sta279-s22.github.io/class_activities/ca_lecture_15.html)
]

---

### Class activity

.large[
.question[
What is the predicted relative risk of short term use vs. no use use for a woman aged 25?
]
]

--

.large[
$\exp\{ -8.234 + 0.456 \cdot 25 - 0.0065 \cdot 25^2 \} = 0.408$
]

---

### Class activity

.large[
.question[
What is the predicted *probability* of each contraceptive choice for a woman aged 25?
]
]

--

.large[
$$\dfrac{\widehat{\pi}_{i(Short)}}{\widehat{\pi}_{i(None)}} = 0.408 \hspace{0.5cm} \Rightarrow \hspace{0.5cm} \widehat{\pi}_{i(Short)} = 0.408 \widehat{\pi}_{i(None)}$$

$$\dfrac{\widehat{\pi}_{i(Long)}}{\widehat{\pi}_{i(None)}} = 1.16 \hspace{0.5cm} \Rightarrow \hspace{0.5cm} \widehat{\pi}_{i(Long)} = 1.16 \widehat{\pi}_{i(None)}$$
]

---

### Class activity

.large[
$$\dfrac{\widehat{\pi}_{i(Short)}}{\widehat{\pi}_{i(None)}} = 0.408 \hspace{0.5cm} \Rightarrow \hspace{0.5cm} \widehat{\pi}_{i(Short)} = 0.408 \widehat{\pi}_{i(None)}$$

$$\dfrac{\widehat{\pi}_{i(Long)}}{\widehat{\pi}_{i(None)}} = 1.16 \hspace{0.5cm} \Rightarrow \hspace{0.5cm} \widehat{\pi}_{i(Long)} = 1.16 \widehat{\pi}_{i(None)}$$

$$\widehat{\pi}_{i(None)} + 0.408 \widehat{\pi}_{i(None)} + 1.16 \widehat{\pi}_{i(None)} = 1$$
$$\widehat{\pi}_{i(None)} = \dfrac{1}{1 + 0.408 + 1.16}$$
]

--

.large[
$$\widehat{\pi}_{i(Short)} = 0.408 \widehat{\pi}_{i(None)} = \dfrac{0.408}{1 + 0.408 + 1.16}$$
$$\widehat{\pi}_{i(Long)} = 1.16 \widehat{\pi}_{i(None)} = \dfrac{1.16}{1 + 0.408 + 1.16}$$
]

---

### Generalizing

.large[
Let $Y_i \sim Categorical(\pi_{i1}, ..., \pi_{iJ})$ be a categorical variable with $J$ levels, and let $j^*$ be the reference level. For each $\pi_{ij}$, $j \neq j^*$, we model

$$\log \left( \dfrac{\pi_{ij}}{\pi_{ij^*}} \right) = \beta_{0(j)} + \beta_{1(j)} X_i$$
Then $\pi_{ij} = \pi_{ij^*} \exp\{ \beta_{0(j)} + \beta_{1(j)} X_i \}$, so

$$\pi_{ij^*} = \dfrac{1}{1 + \sum_{j \neq j^*} \exp\{ \beta_{0(j)} + \beta_{1(j)} X_i \} }$$
$$\pi_{ij} = \dfrac{\exp\{ \beta_{0(j)} + \beta_{1(j)} X_i \}}{1 + \sum_{j \neq j^*} \exp\{ \beta_{0(j)} + \beta_{1(j)} X_i \} }$$
]

---

### Predicted probabilities in R

.large[
We can obtain the predicted probabilities for each individual in the data:

```{r}
probspred <- fitted(m1)
```

Here are the predicted probabilities for the second individual:

```{r}
probspred[2,]
```
]

---

### Making predictions

.large[
```{r}
probspred[2,]
```

This is a 24 year old woman who does not use contraceptives.

.question[
How well did we do at estimating their chances of using contraception?
]
]

--

.large[
The probability is highest for the correct choice, which is good
]

---

### Making predictions

.large[
```{r echo=F}
probspred[2,]
```

This is a 24 year old woman who does not use contraceptives.

.question[
What would our predicted probabilities be if we were just guessing?
]
]

---

### Making predictions

.large[
```{r echo=F}
probspred[2,]
```

This is a 24 year old woman who does not use contraceptives.
]

.large[
* If we don't have any data, our estimated probability would be 1/3 for each level
* If we have data but we don't use age, our estimated probability for each level is just the proportion of observations in that group:

```{r}
table(cmc_data$Choice)/nrow(cmc_data)
```
]

---

### Making predictions

.large[
```{r echo=F}
probspred[2,]
```

This is a 24 year old woman who does not use contraceptives.
]

.large[
.question[
For this individual, are we doing better than random guessing?
]
]

--

.large[
Yes -- our predicted probability for the correct choice (None) is higher than random guessing without data (0.333) and random guessing with data (0.427)
]

---

### Making predictions

.large[
.question[
Given predicted probabilities $\widehat{\pi}_{ij}$ for individual $i$, how could I predict the response $\widehat{Y}_i$?
]
]

--

.large[
* Typically, pick the level with the highest predicted probability
* This is what R does by default
]

---

### Making predictions

.large[
```{r}
probspred <- fitted(m1)
probspred[2,]

preds <- predict(m1)
preds[2]
```
]

---

### Making predictions

.large[
.question[
How can I assess all of my predictions at once?
]
]

--

.large[
With a confusion matrix!

```{r}
table("Prediction" = preds, 
      "Actual" = cmc_data$Choice)
```
]

---

### Making predictions

.large[
```{r echo=F}
table("Prediction" = preds, 
      "Actual" = cmc_data$Choice)
```

.question[
Does it look like we're doing a good job at predicting contraception use?
]
]

---

### Class activity, Part II

.large[
[https://sta279-s22.github.io/class_activities/ca_lecture_15.html](https://sta279-s22.github.io/class_activities/ca_lecture_15.html)
]

---

### Class activity

.large[
```{r echo=F}
table("Prediction" = preds, 
      "Actual" = cmc_data$Choice)
```

.question[
What fraction of our predictions are correct?
]
]

--

.large[
$\dfrac{342 + 0 + 322}{1473} = 0.45$
]

---

### Class activity

.large[
.question[
What would our confusion matrix look like if our predictions randomly assigned each person to one of the three categories, with a 1/3 chance for each category?
]
]

--

.large[
Something like 

| | | Actual | | |
| --- | --- | --- | --- | --- |
| | | None | Short | Long |
|Predicted | None | 210 | 111 | 170 |
| | Short | 210 | 111 | 170 |
| | Long | 209 | 111 | 171 |

]

---

### Class activity

.large[

| | | Actual | | |
| --- | --- | --- | --- | --- |
| | | None | Short | Long |
|Predicted | None | 210 | 111 | 170 |
| | Short | 210 | 111 | 170 |
| | Long | 209 | 111 | 171 |

.question[
What is the accuracy of our predictions in this confusion matrix?
]
]

--

.large[
About $\dfrac{1}{3}$
]

---

### Class activity

.large[
.question[
What would our confusion matrix look like if for every individual, we just predicted the most common contraception choice in the data?
]
]

--

.large[
The most common choice is None, so

| | | Actual | | |
| --- | --- | --- | --- | --- |
| | | None | Short | Long |
|Predicted | None | 629 | 333 | 511 |
| | Short | 0 | 0 | 0 |
| | Long | 0 | 0 | 0 |

]

---

### Class activity

.large[

| | | Actual | | |
| --- | --- | --- | --- | --- |
| | | None | Short | Long |
|Predicted | None | 629 | 333 | 511 |
| | Short | 0 | 0 | 0 |
| | Long | 0 | 0 | 0 |

.question[
What is the accuracy of our predictions in this confusion matrix?
]
]

--

.large[
$\dfrac{629}{1473} = 0.427$ (the prevalence of the most common choice)
]

---

### Class activity

.large[
.question[
Do we do better than random guessing?
]
]

--

.large[
Sort of: 

* We have higher accuracy than either method of random guessing. But our accuracy is only slightly higher than just predicting "None" for everyone
* We do really badly at predicting short-term contraceptive use
]

---

### Moral

.large[
* By itself, accuracy isn't particularly useful for summarizing prediction performance
* We need to look at predictive ability for each class
]