---
title: Simulation
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Exam 2 next Wednesday (April 6)
  * Multinomial regression and mixed effects models
  * Covers up through end of this week
* Today: simulating data
* Wednesday: parametric bootstrap
]

---

### Data and Goal

.large[
```{r include=F}
library(lme4)
library(tidyverse)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

bryozoan <- read_csv("https://sta112-s22.github.io/homework/bryozoan_data.csv")
bugula_early <- bryozoan %>%
  filter(Stage == "early", Species == "bugula")
```

We have date on 197 early-stage *Bugula neritina*, with information on

* `Run`: which repetition of the experiment the individual was recorded in
* `Mass`: Mass of the individual (in micrograms)
* `Metabolic`: Recorded metabolic rate (rate of energy consumption) of the individual (in mJ per hour)

.question[
**Goal for this class:** Is there systematic variation between different runs (i.e., is there any correlation due to Run)?
]
]

---

### Visualizing runs

.large[

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```

.question[
Does it look like there is variation from run to run (i.e., do we need $u_i$ )?
]
]

---

### Visualizing runs

.large[

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```

It is hard to tell if there are actually differences between runs, or if observed differences are just due to chance.

.question[
How can I estimate variability between runs?
]
]

---

### Initial random intercepts model

.large[
$$Metabolic_{ij} = \beta_0 + u_i + \varepsilon_{ij} \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2), \hspace{0.2cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

```{r output.lines=12:19}
m1 <- lmer(Metabolic ~ (1|Run), data = bugula_early)
summary(m1)
```


$\widehat{\beta}_0 = 0.175$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00013$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0042$
]


---

### Initial random intercepts model

.large[
$$Metabolic_{ij} = \beta_0 + u_i + \varepsilon_{ij} \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2), \hspace{0.2cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

$\widehat{\beta}_0 = 0.175$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00013$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0042$

.question[
Does the fitted model suggest there is much variation from run to run?
]

]

--

.large[
No -- the estimated variability between runs is much smaller than the estimated variability within a run.

$\widehat{\rho}_{group} = \dfrac{0.00013}{0.00013 + 0.0042} = 0.03$
]

.large[
.question[
What would the model be *without* the random effect?
]
]

---

### Intercept-only model (no random effect)

.large[
$$Metabolic_{ij} = \beta_0 + \varepsilon_{ij} \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

.question[
How do I fit this in R?
]
]

--

.large[
```{r output.lines=10:15}
m0 <- lm(Metabolic ~ 1, data = bugula_early)
summary(m0)
```

]

---

### Intercept-only model (no random effect)

.large[
$$Metabolic_{ij} = \beta_0 + \varepsilon_{ij} \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

```{r output.lines=10:15}
m0 <- lm(Metabolic ~ 1, data = bugula_early)
summary(m0)
```

$\widehat{\beta}_0 = 0.175, \hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.06558^2 = 0.0043$

]

---

### What we want to know

.large[
Suppose that the intercept-only linear model is correct:
$$Metabolic_{ij} = \beta_0 + \varepsilon_{ij} \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$
How unusual are the observed differences between runs, if there is really no systematic differences between runs (i.e., no random effects)?

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

---

### Plan

.large[
How unusual are the observed differences between runs, if there is really no systematic differences between runs (i.e., no random effects)?

* Pretend that the intercept-only model is correct
* Create a new dataset from the intercept-only model
* Compare our new dataset to the observed dataset
  * If simulated data looks very different, maybe there are systematic differences between runs
  * If simulated data looks similar to observed data, maybe there aren't systematic differences between runs

.question[
**Simulation:** Using the computer to create artificial data that approximates what we might see if we could actually collect new data
]
]

---

### Pretend the intercept-only model is correct

.large[
$$Metabolic_{ij} = \beta_0 + \varepsilon_{ij} \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

.question[
I want to generate a new observation from this relationship. What do I do?
]
]

--

.large[
* Sample a new noise term: $\varepsilon^* \sim N(0, \sigma_\varepsilon^2)$
* Add the intercept: $\beta_0 + \varepsilon^*$

So our new observed metabolic rate is 
$$Metabolic^* = \beta_0 + \varepsilon^*$$
.question[
But, I don't know $\beta_0$ or $\sigma_\varepsilon^2$. What values should I use?
]
]

---

### Pretend the intercept-only model is correct

.large[
$$Metabolic_{ij} = \beta_0 + \varepsilon_{ij} \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

$\widehat{\beta}_0 = 0.175, \hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0043$

* Sample a new noise term: $\varepsilon^* \sim N(0, \ 0.0043)$
* Add the intercept: $0.175 + \varepsilon^*$

So our new observed metabolic rate is 
$$Metabolic^* = 0.175 + \varepsilon^*$$
]

---

### Simulation in R

.large[
Sample a new noise term: $\varepsilon^* \sim N(0, \ 0.0043)$

```{r include=F}
set.seed(3)
```

```{r}
rnorm(n=1, mean=0, sd=sqrt(0.0043))
```

Here $\varepsilon^* = -0.063$

* `rnorm` means "random sample from a normal distribution"
* `n=1` means "the sample size is 1"
* `mean=0` and `sd=sqrt(0.0043)` means "use a $N(0, 0.0043)$ distribution"
]

---

### Simulation in R

.large[
Sample a new noise term: $\varepsilon^* \sim N(0, \ 0.0043)$

```{r include=F}
set.seed(3)
```

```{r}
rnorm(n=1, mean=0, sd=sqrt(0.0043))
```

Here $\varepsilon^* = -0.063$

.question[
Will I get the same $\varepsilon^*$ each time I sample?
]
]

--

.large[
No -- I will get a different $\varepsilon^*$ each time!

.question[
In your R console, run the code to sample a new $\varepsilon^*$ from the normal distribution with mean 0 and variance 0.0043.
]
]

---

### Simulation in R

.large[
* Sample a new noise term: $\varepsilon^* \sim N(0, \ 0.0043)$
* Add the intercept: $0.175 + \varepsilon^*$

```{r include=F}
set.seed(3)
```

```{r}
0.175 + rnorm(n=1, mean=0, sd=sqrt(0.0043))
```

$Metabolic^* = 0.112$

.question[
In your R console, run the code to generate a new metabolic observation $Metabolic^*$.
]
]


---

### Plan (so far)

.large[
$$Metabolic_{ij} = \beta_0 + \varepsilon_{ij} \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$
.question[
How unusual are the observed differences between runs, if there is really no systematic differences between runs (i.e., no random effects)?
]
]

.large[
* Pretend that the intercept-only model is correct
  * $Metabolic^* = 0.175 + \varepsilon^* \hspace{1cm} \varepsilon^* \sim N(0, 0.0043)$
* Create a new dataset from the intercept-only model
* Compare our new dataset to the observed dataset

]

---

### Create a new dataset

.large[
Now we sample a new metabolic rate for **every** observation in the data:

$$Metabolic_{ij}^* = 0.175 + \varepsilon_{ij}^* \hspace{1cm} \varepsilon_{ij}^* \overset{iid}{\sim} N(0, \ 0.0043)$$

```{r}
new_metabolic <- 0.175 +
  rnorm(n=197, mean=0, sd=sqrt(0.0043))
```

.question[
Why do I use `n=197` here?
]
]

--

.large[
There are 197 early-stage *Bugula*
]

---

### Create a new dataset

.large[
$$Metabolic_{ij}^* = 0.175 + \varepsilon_{ij}^* \hspace{1cm} \varepsilon_{ij}^* \overset{iid}{\sim} N(0, \ 0.0043)$$

```{r}
new_metabolic <- 0.175 +
  rnorm(n=197, mean=0, sd=sqrt(0.0043))

new_data <- data.frame(Run = bugula_early$Run,
                       Metabolic = new_metabolic)
```

* Create a new metabolic rate for every organism in the data
* Use the same runs from the observed data
* Store the simulated dataset as `new_data`

]

---

### Create a new dataset

.large[
```{r fig.align='center', fig.width=7, fig.height=5}
new_data %>%
  ggplot(aes(x = as.factor(Run),
             y = Metabolic)) +
  geom_boxplot()
```
]

---

### Plan (so far)

.large[
$$Metabolic_{ij} = \beta_0 + \varepsilon_{ij} \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$
.question[
How unusual are the observed differences between runs, if there is really no systematic differences between runs (i.e., no random effects)?
]
]

.large[
* Pretend that the intercept-only model is correct
  * $Metabolic^* = 0.175 + \varepsilon^* \hspace{1cm} \varepsilon^* \sim N(0, 0.0043)$
* Create a new dataset from the intercept-only model
```{r, eval=F}
new_metabolic <- 0.175 +
  rnorm(n=197, mean=0, sd=sqrt(0.0043))
```
* Compare our new dataset to the observed dataset

]

---

### Class activity, Part I

---

### Class activity

```{r include=F}
music <- read_csv("https://sta279-s22.github.io/labs/music.csv")
```

.large[
```{r, output.lines=12:15}
m1 <- lmer(na ~ (1|id), data = music)
summary(m1)
```

.question[
What is the estimated intra-class correlation?
]
]

--

.large[
$\widehat{\rho}_{group} = \dfrac{4.95}{4.95 + 22.46} = 0.18$
]

---

### Class activity

.large[
```{r output.lines=10:15}
m0 <- lm(na ~ 1, data = music)
summary(m0)
```

.question[
What are the estimates $\widehat{\beta}_0$ and $\widehat{\sigma}_\varepsilon^2$?
]
]

--

.large[
$\widehat{\beta}_0 = 16.21$, $\widehat{\sigma}_\varepsilon^2 = 5.237^2 = 27.43$
]

---

### Class activity

.large[
$$Anxiety_{ij}^* = \widehat{\beta}_0 + \varepsilon_{ij}^* \hspace{1cm} \varepsilon_{ij}^* \sim N(0, \widehat{\sigma}_\varepsilon^2)$$

```{r eval=F}
new_na <- ... +
  rnorm(n=..., mean=0, sd=...)

new_data <- data.frame(id = music$id,
                       na = new_na)
```

.question[
How do I fill in the code to simulate a new dataset from the intercept-only model?
]

]

---

### Class activity

.large[
$$Anxiety_{ij}^* = \widehat{\beta}_0 + \varepsilon_{ij}^* \hspace{1cm} \varepsilon_{ij}^* \sim N(0, \widehat{\sigma}_\varepsilon^2)$$

$\widehat{\beta}_0 = 16.21$, $\widehat{\sigma}_\varepsilon^2 = 5.237^2 = 27.43$

```{r eval=F}
new_na <- 16.21 +
  rnorm(n=497, mean=0, sd=5.237)

new_data <- data.frame(id = music$id,
                       na = new_na)
```

]

---

### Compare new dataset to observed dataset

.pull-left[
.large[
**Original (observed) data:**
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
**New (simulated) data:**
]
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
new_data %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.large[
.question[
Do you think there is systematic variation between runs, or do you think the observed differences between runs are due to chance?
]
]

---

### Compare new dataset to observed dataset

.pull-left[
.large[
**Original (observed) data:**
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
**New (simulated) data:**
]
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
new_data %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.large[
* There seems to be similar differences between runs for the observed and simulated data
* We know there is no random effect in the simulated data
* So maybe we don't need a random effect for the observed data?
]

---

### Compare new dataset to observed dataset

.large[
$$Metabolic_{ij} = \beta_0 + u_i + \varepsilon_{ij} \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2), \hspace{0.2cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

**Fitted random intercepts model (observed data):**

$\widehat{\beta}_0 = 0.175$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00013$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0042$

**Fitted random intercepts model (simulated data):**

$\widehat{\beta}_0 = 0.169$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00015$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0049$

.question[
Do you think there is systematic variation between runs, or do you think the observed differences between runs are due to chance?
]
]

---

### Compare new dataset to observed dataset

.large[
$$Metabolic_{ij} = \beta_0 + u_i + \varepsilon_{ij} \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2), \hspace{0.2cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

**Fitted random intercepts model (observed data):**

$\widehat{\beta}_0 = 0.175$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00013$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0042$

**Fitted random intercepts model (simulated data):**

$\widehat{\beta}_0 = 0.169$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00015$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0049$

* Estimated variances are similar for observed data and simulated data
* We know there is no random effect in the simulated data
* So maybe observed data isn't unusual if there are no systematic differences between runs
]

---

### Summary (so far)

.large[
.question[
Are there systematic differences between runs (group effects), or are observed differences simply due to chance?
]

* Fit a model with no random effects
* Simulate data from fitted model
* Compare simulated data to observed data
  * If simulated data looks very different, maybe there are systematic differences between runs
  * If simulated data looks similar to observed data, maybe there aren't systematic differences between runs
]

---

### Class activity, Part II

