---
title: Random slopes
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Recap random slopes
* Fitting and interpreting mixed effects models with random slopes
* Lab 7
]

---

### Adding random slopes

.large[
$$Price_{ij} = \beta_0 + u_i + (\beta_1 + v_i) Satisfaction_{ij} + \varepsilon_{ij}$$

* $\beta_0$ = mean price when satisfaction is 0 (average across neighborhoods)
* $\beta_0 + u_i$ = mean price when satisfaction is 0 in neighborhood $i$
* $\beta_1$ = average change in price for a one-unit increase in satisfaction (average across neighborhoods)
* $\beta_1 + v_i$ = average change in price for a one-unit increase in satisfaction in neighborhood $i$
]

---

### Example: music performance anxiety

.large[
**Data:** Pre-performance anxiety levels for 37 undergraduate music majors (between 2 and 15 performances for each student). Each row in the data represents one performance, and variables include:

* `id`: a unique identifier for the musician
* `na`: negative affect score (a measure of anxiety)
* `large`: whether the musician was performing as part of a large ensemble (large = 1), or as part of a small ensemble or solo (large = 0)

.question[
What are the groups in the data, and what are the observations within each group?
]
]

---

### Example: music performance anxiety

.large[
.question[
What are the groups in the data, and what are the observations within each group?
]

* The groups are individual musicians
* The observations are performances for each musician
]

---

### Example: music performance anxiety

.large[
Researchers hypothesize that:
* overall anxiety varies from student to student
* anxiety will be lower before large ensemble performance
* the difference in performance anxiety between large and small performances will vary from student to student

.question[
What mixed effects model should the researchers use to investigate their hypothesis?
]
]

--

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

where $Anxiety_{ij}$ is the performance anxiety of musician $i$ before performance $j$.
]

---

### Example: music performance anxiety

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

.question[
Interpret the fixed effects and random effects in the model.
]
]

--

.large[
* $\beta_0$ = average performance anxiety before small ensemble and solo performances (average across musicians)
* $\beta_0 + u_i$ = average performance anxiety before small ensemble and solo performances for musician $i$
* $\beta_1$ = average difference in anxiety before large ensemble performances (compared to small/solo performances) (average across musicians)
* $\beta_1 + v_i$ = average difference in anxiety before large ensemble performances for musician $i$
]

---

### Fitting a model with random slopes

.large[
```{r, eval=F}
m1 <- lmer(na ~ large + (large|id),
           data = music)
```

* This means we include `large` as a fixed effect, *and* we allow the coefficient on `large` to vary from individual to individual
]

---

### Fitting a model with random slopes

```{r include=F}
library(tidyverse)
library(lme4)
library(knitr)

music <- read.csv("musicdata.csv") %>%
   mutate(large = ifelse(perform_type == "Large Ensemble", 1, 0))

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

.large[
```{r, output.lines = 18:22}
m1 <- lmer(na ~ large + (large|id),
           data = music)
summary(m1)
```

$\widehat{\beta}_0 = 16.73 \hspace{1cm} \widehat{\beta}_1 = -1.68$

.question[
How would I interpret $\widehat{\beta}_0$ and $\widehat{\beta}_1$?
]
]

---

### Fitting a model with random slopes

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

$\widehat{\beta}_0 = 16.73 \hspace{1cm} \widehat{\beta}_1 = -1.68$

* The average anxiety before a small/solo performance is 16.73
* On average, student anxiety decreases by 1.68 before a large performance (compared to a small/solo performance)
]

---

### Fitting a model with random slopes

.large[
```{r, output.lines = 11:16}
summary(m1)
```

.question[
What does this output tell us about the random effects and the noise?
]
]

---

### Fitting a model with random slopes


.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

```{r, echo=F, output.lines = 11:15}
summary(m1)
```

* $\widehat{\sigma}_u^2 = 6.333$ (variability in anxiety before small performances, between students)
* $\widehat{\sigma}_v^2 = 0.743$ (variability in difference in anxiety before large performances, between students)
* $\widehat{\sigma}_\varepsilon^2 = 21.77$ (variability in anxiety between performances, within a student)
]

---

### Correlation between slopes and intercepts

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$

```{r, echo=F, output.lines = 11:15}
summary(m1)
```

* $\widehat{\rho}_{uv} = -0.76$ (estimated correlation between the random slope and random intercept for an individual)
  * Subjects with higher performance anxiety scores for solos and small ensembles tend to have greater decreases in performance anxiety for large ensemble performances
]

---

### Writing down the model

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$
$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} \begin{bmatrix} u_i \\ v_i \end{bmatrix} \overset{iid}{\sim} N \left( \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix} \sigma_u^2 & \rho_{uv}\sigma_u \sigma_v \\ \rho_{uv}\sigma_u \sigma_v & \sigma_v^2 \end{bmatrix} \right)$

.question[
Anybody know the name of this new thing?
]
]

--

.large[
A bivariate normal distribution!
]

---

### Writing down the model

.large[
$$Anxiety_{ij} = \beta_0 + u_i + (\beta_1 + v_i) LargeEnsemble_{ij} + \varepsilon_{ij}$$
$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} \begin{bmatrix} u_i \\ v_i \end{bmatrix} \overset{iid}{\sim} N \left( \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix} \sigma_u^2 & \rho_{uv}\sigma_u \sigma_v \\ \rho_{uv}\sigma_u \sigma_v & \sigma_v^2 \end{bmatrix} \right)$

* This just says that both $u_i$ and $v_i$ come from a normal distribution
  * the variance of $u_i$ is $\sigma_u^2$
  * the variance of $v_i$ is $\sigma_v^2$
  * the correlation between $u_i$ and $v_i$ is $\rho_{uv}$
* *Note:* the population model includes the distribution of the random effects and noise

]