---
title: Logistic regression with multiple predictors
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Lab 2 due tomorrow (Tuesday)
* Exam 1 moved to Monday, February 14
* Today:
  * multiple logistic regression
  * likelihood ratio tests
  * AIC
]

---

### Scenario: dengue fever in Vietnam

.large[
* *Dengue fever*: a mosquito-borne viral disease, which infects hundreds of millions of people a year. Common in tropical climates
* Researchers in Vietnam are interested in identifying the relationship between specific symptoms and the probability of having dengue
  * E.g.: a patient goes to the hospital with fever and abdominal pain. What is the probability they have dengue?
* Lab tests are available, but may take time to return results and/or be expensive
]

---

### Data

.large[
Data on 5720 Vietnamese children, admitted to hospital with possible dengue fever. Variables include:
* `Dengue`: whether the patient actually has dengue fever, based on a lab test (0 = no, 1 = yes)
* `Temperature`: patient's body temperature (in Celsius)
* `Abdominal`: whether the patient has abdominal pain (0 = no, 1 = yes)
* `HCT`: patient's hematocrit (proportion of red blood cells)
* `Age`: patient's age (in years)
* `Sex`: patient's sex 
* `+` several others
]

---

### Research question

.large[
You are approached by the researchers to help analyze their data. Their initial question: 

*What is the relationship between temperature and the probability of dengue, and does the relationship differ if the patient also presents with abdominal pain?*
]

---

### Exploratory data analysis

.large[
.question[
How can we visualize the relationship between temperature and the probability of having dengue?
]
]

--

.large[
*An empirical log odds plot!*
]

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=5, fig.align='center'}
library(tidyverse)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       #x <- c(head(x, lines), more)
       x <- c(head(x, lines))
     }
   } else {
     #x <- c(more, x[lines], more)
     x <- c(x[lines])
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

dengue_original <- read.csv("dengue_original.csv")

logodds_plot <- function(data, num_bins, bin_method,
                         x, y, grouping = NULL, reg_formula = y ~ x){
  
  if(is.null(grouping)){
    dat <- data.frame(x = data[,x], 
                      y = data[,y],
                      group = 1)
  } else {
    dat <- data.frame(x = data[,x], 
               y = data[,y],
               group = data[,grouping])
  }
  
  if(bin_method == "equal_size"){
    logodds_table <- dat %>%
      drop_na() %>%
      arrange(group, x) %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = rep(1:num_bins,
                       each=ceiling(n()/num_bins))[1:n()]) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  } else {
    logodds_table <- dat %>%
      drop_na() %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = cut(x, 
                       breaks = num_bins,
                       labels = F)) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  }
  
  if(is.null(grouping)){
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds)) +
      geom_point(size=2.5) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x,
           y = "Empirical log odds") +
      theme(text = element_text(size=25))
  } else {
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds,
                 color = group,
                 shape = group)) +
      geom_point(size=2.5) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x,
           y = "Empirical log odds",
           color = grouping,
           shape = grouping) +
      theme(text = element_text(size=25))
  }
  
}

### Do some data cleaning
dengue <- dengue_original %>%
  select(SiteNo, Sex, Age, DayDisease, Vomiting, Abdo, Muco, Skin,
         Temp, BMI, Height, Weight, Flush, Rash, WBC, HCT, PLT, 
         NS1_TRIP, Lab_Confirmed_Dengue) %>%
  rename(SiteNumber = SiteNo, 
         DiseaseDay = DayDisease,
         Abdominal = Abdo,
         Mucosal = Muco,
         Temperature = Temp, 
         RapidTest = NS1_TRIP,
         Dengue = Lab_Confirmed_Dengue) %>%
  mutate(Vomiting = as.factor(2 - Vomiting),
         Abdominal = as.factor(2 - Abdominal),
         Mucosal = as.factor(2 - Mucosal),
         Skin = as.factor(2 - Skin),
         Flush = as.factor(2 - Flush),
         Rash = as.factor(2 - Rash),
         RapidTest = ifelse(RapidTest == 1, "positive", "negative"),
         Dengue = 2 - Dengue,
         Sex = ifelse(Sex == 1, "male", "female")) %>% 
  drop_na()

logodds_plot(dengue, 12, "equal_size", x = "Temperature", y = "Dengue")
```

.large[
.question[
Based on the empirical log odds plot, what would be a reasonable model for the relationship between temperature and the probability a patient has dengue?
]
]

---

### Initial model

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 \ Temperature_i$$
]

--

.large[
We're also interested in whether this relationship is different depending on abdominal pain.
.question[
What plot can I use to investigate this question?
]
]

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
logodds_plot(dengue, 10, "equal_size", x = "Temperature", y = "Dengue",
             grouping = "Abdominal")
```

.large[
.question[
Based on this plot, what would be a reasonable model that incorporates both temperature and abdominal pain?
]
]

---

### Adding abdominal pain

.large[
$$Y_i \sim Bernoulli(\pi_i)$$
]

.large[
\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ & \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
* The interaction term allows the relationship between temperature and $\pi_i$ to change depending on abdominal pain
]

---

### Class activity, Part I

.large[
[https://sta279-s22.github.io/class_activities_ca_lecture_10.html](https://sta279-s22.github.io/class_activities_ca_lecture_10.html)
]

---

### Class activity

.large[
\begin{align}
\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) &= -7.745 + 0.178 \ Temperature_i \\
& \hspace{0.5cm} - 6.129 \ Abdominal_i \\ 
& \hspace{0.5cm} + 0.166 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
.question[
What is the estimated probability of dengue for a patient with a temperature of 38C and abdominal pain?
]
]

--

.large[
\begin{align}
\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) &= -7.745 + 0.178(38) - 6.129 + 0.166(38) \\
&= -0.802 \\
\widehat{\pi}_i &= 0.31
\end{align}
]

---

### Class activity

.large[
\begin{align}
\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) &= -7.745 + 0.178 \ Temperature_i \\
& \hspace{0.5cm} - 6.129 \ Abdominal_i \\ 
& \hspace{0.5cm} + 0.166 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
.question[
For patients with abdominal pain, what is the estimated change in odds associated with an increase in temperature of 1C?
]
]

--

.large[
$\widehat{\beta}_1 + \widehat{\beta}_3 = 0.344$

$e^{0.344} = 1.41$

So an increase in temperature of 1C is associated with an increase in the odds of dengue by a factor of 1.41.
]

---

### Class activity

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
I want to test whether there is a relationship between abdominal pain and the probability of dengue, after accounting for the relationship between temperature and the probability of dengue.
.question[
What are my null and alternative hypotheses?
]
]

--

.large[
$$H_0: \beta_2 = \beta_3 = 0$$

$$H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$
]

---

### Hypothesis testing

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
**Hypotheses:**

$$H_0: \beta_2 = \beta_3 = 0$$

$$H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$
]

.large[
.question[
Which type of test can I use to test these hypotheses ( Wald test, likelihood ratio test, or either)?
]
]

---

### Hypothesis testing

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
**Hypotheses:**

$$H_0: \beta_2 = \beta_3 = 0$$

$$H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$
]

.large[
We will use a likelihood ratio test, since we are testing multiple parameters at once.
]

---

### Likelihood ratio test

.large[
$$Y_i \sim Bernoulli(\pi_i)$$

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}
]

.large[
**Hypotheses:**

$$H_0: \beta_2 = \beta_3 = 0$$

$$H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$
]

.large[
.question[
What are my full and reduced models?
]
]

---

### Full and reduced models

.large[
**Full model:**

\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\ 
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i
\end{align}

**Reduced model:**

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 \ Temperature_i$$
]

---

### Test statistic

.large[
$$G = \text{deviance of full model } - \text{deviance of reduced model}$$
]

.large[
**Full model:**

```{r echo=F, output.lines = 12:13}
glm(Dengue ~ Temperature*Abdominal, family=binomial, data=dengue)
```
]

.large[
**Reduced model:**

```{r echo=F, output.lines = 9:10}
glm(Dengue ~ Temperature, family=binomial, data=dengue)
```
]

.large[
$$G = 6927 - 6914 = 13$$
.question[
How do I calculate a p-value for this test statistic?
]
]

---

### p-value

.large[
$$G = 6927 - 6914 = 13 \hspace{1cm} G \sim \chi^2_k$$
where $k =$ difference in number of parameters between full and reduced models.

.question[
What is $k$ for this test?
]
]

--

.large[
$H_0: \beta_2 = \beta_3 = 0$, so $k = 2$

```{r}
pchisq(13, df=2, lower.tail=F)
```
]

---

### Conclusion

.large[
* **Question:** Is there a relationship between abdominal pain and the probability of dengue, after accounting for the relationship between temperature and the probability of dengue?
* **Hypotheses:**
$$H_0: \beta_2 = \beta_3 = 0$$

$$H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$
* **p-value:** 0.0015
* **Conclusion:** There is strong evidence that there is a relationship between abdominal pain and the probability of dengue, after accounting for the relationship between temperature and the probability of dengue.
]

---

### A new question...

.large[
You report your results to the hospital, and they ask a follow-up question: 

*Does the model improve when we add hematocrit (the proportion of red blood cells)?*
]

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
logodds_plot(dengue, 10, "equal_size", x = "HCT", y = "Dengue",
             grouping = "Abdominal")
```

.large[
.question[
Does it look like we need an interaction between hematocrit (HCT) and abdominal pain?
]
]

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
logodds_plot(dengue, 10, "equal_size", x = "HCT", y = "Dengue",
             grouping = "Abdominal")
```

.large[
.question[
How can I check whether there might be an interaction between temperature and hematocrit?
]
]

---

### Exploratory data analysis

.large[
Define $TempHigh$ by 
* $TempHigh_i = 1$ if $Temperature_i > 38$
* $TempHigh_i = 0$ if $Temperature_i <= 38$
]

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
dengue %>%
  mutate(TempHigh = as.factor((Temperature > 38) + 0)) %>%
  logodds_plot(10, "equal_size", x = "HCT", y = "Dengue",
             grouping = "TempHigh")
```

---

### Exploratory data analysis

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=5, fig.align='center'}
dengue %>%
  mutate(TempHigh = as.factor((Temperature > 38) + 0)) %>%
  logodds_plot(10, "equal_size", x = "HCT", y = "Dengue",
             grouping = "TempHigh")
```

.large[
.question[
Does it look like we need an interaction between temperature and hematocrit?
]
]

---

### Model

.large[
$$Y_i \sim Bernoulli(\pi_i)$$
]

.large[
\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i \\
& \hspace{0.5cm} + \beta_4 \ HCT_i
\end{align}
]

.large[
* Note that while we binarized temperature for EDA, we use the original variable in the model here
]

---

### Class activity, Part II

.large[
[https://sta279-s22.github.io/class_activities_ca_lecture_10.html](https://sta279-s22.github.io/class_activities_ca_lecture_10.html)
]

---

### Class activity

.large[
.question[
What is the estimated change in odds associated with a 1 point increase in hematocrit, holding temperature and abdominal pain constant?
]
]

--

.large[
$\widehat{\beta}_4 = 0.115$

So we estimate that a 1 point increase in hematocrit is associated with an increase in the odds of dengue by a factor of $e^{0.115} = 1.12$
]

---

### Class activity

.large[
.question[
How does the deviance change when we add hematocrit to the model?
]
]

--

.large[
It will decrease -- we can always decrease the deviance by adding more terms.

Indeed, the deviance decreases from 6914 to 6745.
]

---

### Comparing models

.large[
If deviance always decreases when I add additional variables, how can I assess whether including hematocrit substantially improves the model?

**Option 1:** Likelihood ratio test
* Is the change in deviance bigger than we would expect if hematocrit doesn't really matter?
]

---

### Likelihood ratio test

.large[
\begin{align}
\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) &= \beta_0 + \beta_1 \ Temperature_i + \beta_2 \ Abdominal_i \\
& \hspace{0.5cm} + \beta_3 \ Temperature_i \cdot Abdominal_i \\
& \hspace{0.5cm} + \beta_4 \ HCT_i
\end{align}
]

.large[
**Hypotheses:** 

$$H_0: \beta_4 = 0 \hspace{1cm} H_A: \beta_4 \neq 0$$
**Test statistic:**

$$G = \text{deviance for reduced model } - \text{deviance for full model}$$
]

---

### Likelihood ratio test

.large[
$$G = 6914 - 6745 = 169 \hspace{1cm} G \sim \chi^2_k$$

$k = \ ?$
]

---

### Likelihood ratio test

.large[
$$G = 6914 - 6745 = 169 \hspace{1cm} G \sim \chi^2_k$$

$k =$ difference in number of parameters $= 1$

```{r}
pchisq(169, df=1, lower.tail = F)
```
]

---

### Comparing models

.large[
If deviance always decreases when I add additional variables, how can I assess whether including hematocrit substantially improves the model?

**Option 1:** Likelihood ratio test
* Is the change in deviance bigger than we would expect if hematocrit doesn't really matter?

**Option 2:** AIC
]

---

### AIC

.large[
.question[
In linear regression, what quantity did we use to compare models with different numbers of parameters?
]
]

--

.large[
*Adjusted $R^2$*

* We can use something similar for logistic regression, called the *Akaike information criterion* (AIC)
* Motivation: penalize the deviance based on the number of parameters
]

---

### AIC

.large[
**AIC:** Suppose our model has $p$ parameters (including the intercept). Then the AIC is
$$AIC = 2p + \text{deviance}$$
]

.large[
* Penalize deviance based on the number of parameters
* Is the reduction in deviance worth adding more parameters to the model?
* We prefer a smaller AIC
]

---

### AIC

.large[
**Model 1:** (adding hematocrit)

```{r echo=F, output.lines = 12:13}
glm(Dengue ~ Temperature*Abdominal + HCT, family=binomial, data=dengue)
```
]

.large[
**Model 2:** (no hematocrit)

```{r echo=F, output.lines = 12:13}
glm(Dengue ~ Temperature*Abdominal, family=binomial, data=dengue)
```
]

.large[
.question[
Which model do we prefer, based on AIC?
]
]

--

.large[
Model 2
]

---

### Model comparison

.large[
*Does the model improve when we add hematocrit (the proportion of red blood cells)?*

* **Likelihood ratio test:** p-value $\approx 0$
* **AIC:** AIC is smaller when we add hematocrit

**Conclusion:** We have convincing evidence that adding hematocrit improves the model.
]

---

### Next time...

.large[
* prediction
* model selection
]