---
title: Multinomial logistic regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Project 1 due Friday
* Today: multinomial logistic regression
]

---

### Motivation

.large[
* When the response is binary, we use *logistic regression*
* What happens when the response is categorical, but has MORE than 2 categories?
* We use *multinomial logistic regression* (aka *multinomial regression*)
]

---

### Motivation

.large[
**Question:** What is the relationship between age and contraceptive use for women in Indonesia?

**Data:** 1473 Indonesian couples, with variables
* $Y_i$ = contraceptive method used (1 = no use, 2 = long-term, 3 = short-term)
* $X_i$ = Wife's age (numeric)
]

---

### The response variable

.large[
```{r echo=F, message=F, warning=F}
library(tidyverse)
cmc_data <- read_csv("cmc.data.txt", col_names = F)
colnames(cmc_data) <-c("WifeAge", "WifeEdu", "HusbandEdu", "NumChild", "WifeRel", "WifeWork", "HusbandOcc","Living", "Media", "Choice")
cmc_data$Choice <- ifelse( cmc_data$Choice == 1, "None", ifelse(cmc_data$Choice==2, "Short", "Long"))
knitr::kable(table("Contraception" = cmc_data$Choice))
```

* $n_{None} = 629$ (this is 42.7% of the couples)
* $n_{Long} = 511$ (this is 34.7% of the couple)
* $n_{Short} = 333$ (this is 22.6% of the couples)
]

---

### The response variable

.large[
$Y_i$ = contraceptive method used (1 = no use, 2 = long-term, 3 = short-term)

.question[
What type of variable is $Y$?
]
]

--

.large[
Categorical, with three levels. There is no inherent order to the levels.
]

.large[
.question[
Can we use logistic regression?
]
]

--

.large[
No -- $Y_i$ does not come from a Bernoulli distribution.
]

---

### Parametric model building

.large[
**Step 1: Choose a reasonable distribution for $Y$**

* But we don't know any distributions for a categorical variable with more than two levels

**Step 2: Choose a model for any parameters**

* We need Step 1 first!
* Let's learn a new distribution for categorical variables with more than 2 levels
]

---

### Building a distribution

.large[
$Y_i$ = contraceptive method used (1 = no use, 2 = long-term, 3 = short-term)

.question[
What notation might we use for the probability of no contraceptive use?
]
]

--

.large[
$\pi_{i(None)}$

.question[
What about the probabilities of short term and long term use?
]
]

--

.large[
$\pi_{i(Short)}$ and $\pi_{i(Long)}$
]

---

### Building a distribution

.large[
$Y_i$ = contraceptive method used (1 = no use, 2 = long-term, 3 = short-term)

* $\pi_{i(None)} = P(Y_i = None)$
* $\pi_{i(Short)} = P(Y_i = Short)$
* $\pi_{i(Long)} = P(Y_i = Long)$

.question[
What must be true of the three probabilities?
]
]

--

.large[
* They are each between 0 and 1
* They add up to 1: $\pi_{i(None)} + \pi_{i(Short)} + \pi_{i(Long)} = 1$
]

---

### The Categorical distribution

.large[
.question[
**Definition:** Let $Y_i$ be an **unordered** categorical variable with $J$ levels $j = 1,...,J$. Let $\pi_j = P(Y_i = j)$, where $\pi_j \in [0, 1]$ for all $j$, and $\sum \limits_{j=1}^J \pi_j = 1$.

Then we say $Y_i \sim Categorical(\pi_1,...,\pi_J)$.
]

* We can use this distribution as the first step in our modeling process!

.question[
What distribution does our response (contraceptive use) have?
]
]

---

### Parametric model building

.large[
**Step 1: Choose a reasonable distribution for $Y$**

$$Y_i \sim Categorical(\pi_{i(None)}, \pi_{i(Short)}, \pi_{i(Long)})$$

**Step 2: Choose a model for any parameters**

* Need to relate our probabilities to $X = Age$
]

---

### EDA

.large[
```{r echo=F, message=F, fig.align='center', fig.width=7, fig.height=5}
cmc_data %>%
  ggplot(aes(x = Choice, y = WifeAge, fill = Choice)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Contraceptive choice", y = "Wife's Age (years)") +
  theme(text = element_text(size=25))
```
]

---

### EDA

.large[
* Boxplots show there may be some differences with age, but don't let us model the relationship
* We want something like an empirical log odds plot

.question[
Can we use the log odds here?
]
]

--

.large[
Sort of -- instead of odds, we will use the *relative risk*
]

---

### Relative risk

.large[
* If $Y_i$ is *binary*, the odds $\dfrac{\pi_i}{1 - \pi_i}$ compare the probabilities of the two possible outcomes
* If $Y_i$ has more than two outcomes, we need to generalize the odds
* The *relative risk* compares the probabilities of two potential outcomes

**Relative risk of long term vs. no contraceptive use:**

$$\dfrac{\pi_{i(Long)}}{\pi_{i(None)}}$$
]

---

### Relative risk

.large[
* If $Y_i$ is *binary*, the odds $\dfrac{\pi_i}{1 - \pi_i}$ compare the probabilities of the two possible outcomes
* If $Y_i$ has more than two outcomes, we need to generalize the odds
* The *relative risk* compares the probabilities of two potential outcomes

**Relative risk of short term vs. no contraceptive use:**

$$\dfrac{\pi_{i(Short)}}{\pi_{i(None)}}$$
]

---

### Example

.large[
Consider the 48 twenty-one year old wives in our data:

* Long: 23
* Short: 6
* None: 19

.question[
For a 21 year old, what is the *empirical* relative risk of using long term vs. short term contraceptives?
]
]

--

.large[
$\dfrac{23/48}{6/48} = \dfrac{23}{6} = 3.83$
]

---

### Relative risk

.large[
.question[
**Definition:** Let $Y_i$ be a categorical variable with $J$ levels $j=1,...,J$. Let $\pi_j = P(Y_i = j)$. Then the relative risk of level $j$ vs. level $k$ is 

$$\dfrac{\pi_{ij}}{\pi_{ik}}$$
]
]

---

### Class activity, Part I

---

### Log relative risk

.large[
Instead of modeling the log odds, we can model the *log relative risk*
]

```{r echo=F, fig.width=7, fig.height=5, fig.align='center'}
cmc_data %>%
  group_by(WifeAge) %>%
  summarize(count_long = sum(Choice == "Long"),
            count_none = sum(Choice == "None")) %>%
  mutate(log_rr = log(count_long/count_none)) %>%
  ggplot(aes(x = WifeAge, y = log_rr)) +
  geom_point(size=2) +
  labs(x = "Wife's Age", y = "Log RR: Long vs. None") +
  theme_bw() +
  theme(text = element_text(size=25))
```

---

### Log relative risk

.large[
Instead of modeling the log odds, we can model the *log relative risk*
]

```{r echo=F, fig.width=7, fig.height=5, fig.align='center'}
cmc_data %>%
  group_by(WifeAge) %>%
  summarize(count_long = sum(Choice == "Short"),
            count_none = sum(Choice == "None")) %>%
  mutate(log_rr = log(count_long/count_none)) %>%
  ggplot(aes(x = WifeAge, y = log_rr)) +
  geom_point(size=2) +
  labs(x = "Wife's Age", y = "Log RR: Short vs. None") +
  theme_bw() +
  theme(text = element_text(size=25))
```

---

### Multinomial regression model

.large[
**Step 1: Choose a reasonable distribution for $Y$**

$$Y_i \sim Categorical(\pi_{i(None)}, \pi_{i(Short)}, \pi_{i(Long)})$$
**Step 2: Choose a model for any parameters**

$$\log \left( \dfrac{\pi_{i(Long)}}{\pi_{i(None)}} \right) = \beta_{0(Long)} + \beta_{1(Long)} Age_i$$
$$\log \left( \dfrac{\pi_{i(Short)}}{\pi_{i(None)}} \right) = \beta_{0(Short)} + \beta_{1(Short)} Age_i$$
* Pick a *reference* or *baseline* category to compare to (here it is None)
]

---

### Multinomial regression model

.large[
**Step 1: Choose a reasonable distribution for $Y$**

$$Y_i \sim Categorical(\pi_{i(None)}, \pi_{i(Short)}, \pi_{i(Long)})$$
**Step 2: Choose a model for any parameters**

$$\log \left( \dfrac{\pi_{i(Long)}}{\pi_{i(None)}} \right) = \beta_{0(Long)} + \beta_{1(Long)} Age_i$$
$$\log \left( \dfrac{\pi_{i(Short)}}{\pi_{i(None)}} \right) = \beta_{0(Short)} + \beta_{1(Short)} Age_i$$

.question[
From the empirical log relative risk plots, did it look like the log relative risk was a linear function of Age?
]
]

---

### Log relative risk

```{r echo=F, fig.width=7, fig.height=5, fig.align='center'}
cmc_data %>%
  group_by(WifeAge) %>%
  summarize(count_long = sum(Choice == "Long"),
            count_none = sum(Choice == "None")) %>%
  mutate(log_rr = log(count_long/count_none)) %>%
  ggplot(aes(x = WifeAge, y = log_rr)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm", formula= y ~ poly(x,2)) +
  labs(x = "Wife's Age", y = "Log RR: Long vs. None") +
  theme_bw() +
  theme(text = element_text(size=25))
```

---

### Multinomial regression model

.large[
**Step 1: Choose a reasonable distribution for $Y$**

$$Y_i \sim Categorical(\pi_{i(None)}, \pi_{i(Short)}, \pi_{i(Long)})$$
**Step 2: Choose a model for any parameters**

$$\log \left( \dfrac{\pi_{i(Long)}}{\pi_{i(None)}} \right) = \beta_{0(Long)} + \beta_{1(Long)} Age_i + \beta_{2(Long)} Age_i^2$$
$$\log \left( \dfrac{\pi_{i(Short)}}{\pi_{i(None)}} \right) = \beta_{0(Short)} + \beta_{1(Short)} Age_i + \beta_{2(Short)} Age_i^2$$
]

---

### Estimated model

.large[
$$\log \left( \dfrac{\widehat{\pi}_{i(Long)}}{\widehat{\pi}_{i(None)}} \right) = -5.07 + 0.37 Age_i - 0.0063 Age_i^2$$

$$\log \left( \dfrac{\widehat{\pi}_{i(Short)}}{\widehat{\pi}_{i(None)}} \right) = -8.21 + 0.46 Age_i - 0.0065 Age_i^2$$

.question[
What is the predicted relative risk of long term vs. none for a woman age 30?
]
]

--

.large[
$\dfrac{\widehat{\pi}_{i(Long)}}{\widehat{\pi}_{i(None)}} = \exp \{ -5.07 + 0.37(30) - 0.0063(30^2) \} = 1.433$
]

---

### Class activity, Part II