---
title: The Parametric Bootstrap 
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Reminder: Lab 7 due Friday
* Today: 
  * Recap: simulation
  * parametric bootstrap
]

---

### Data and Goal

.large[

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
library(lme4)
library(tidyverse)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

bryozoan <- read_csv("https://sta112-s22.github.io/homework/bryozoan_data.csv")
bugula_early <- bryozoan %>%
  filter(Stage == "early", Species == "bugula")

bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```

.question[
Are observed differences in runs due to systematic variation between runs, or just random chance?
]
]

---

### Plan (so far)

.large[
$$Metabolic_{ij} = \beta_0 + \varepsilon_{ij} \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$
.question[
How unusual are the observed differences between runs, if there is really no systematic differences between runs (i.e., no random effects)?
]
]

.large[
* Pretend that the intercept-only model is correct
  * $Metabolic^* = 0.175 + \varepsilon^* \hspace{1cm} \varepsilon^* \sim N(0, 0.0043)$
* Create a new dataset from the intercept-only model
```{r, eval=F}
new_metabolic <- 0.175 +
  rnorm(n=197, mean=0, sd=sqrt(0.0043))
```
* Compare our new dataset to the observed dataset

]

---

### Our simulated data

.large[
$$Metabolic_{ij}^* = 0.175 + \varepsilon_{ij}^* \hspace{1cm} \varepsilon_{ij}^* \overset{iid}{\sim} N(0, \ 0.0043)$$

```{r}
new_metabolic <- 0.175 +
  rnorm(n=197, mean=0, sd=sqrt(0.0043))

new_data <- data.frame(Run = bugula_early$Run,
                       Metabolic = new_metabolic)
```

* Create a new metabolic rate for every organism in the data
* Use the same runs from the observed data
* Store the simulated dataset as `new_data`

]

---

### Compare new dataset to observed dataset

.pull-left[
.large[
**Original (observed) data:**
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
**New (simulated) data:**
]
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
new_data %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.large[
.question[
Do you think there is systematic variation between runs, or do you think the observed differences between runs are due to chance?
]
]

---

### Compare new dataset to observed dataset

.pull-left[
.large[
**Original (observed) data:**
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
**New (simulated) data:**
]
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
new_data %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.large[
* There seems to be similar differences between runs for the observed and simulated data
* We know there is no random effect in the simulated data
* So maybe we don't need a random effect for the observed data?
]

---

### Compare new dataset to observed dataset

.pull-left[
.large[
**Original (observed) data:**
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
bugula_early %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
**New (simulated) data:**
]
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
new_data %>%
  ggplot(aes(x = as.factor(Run), y = Metabolic)) +
  geom_boxplot(lwd=1) +
  theme_bw() +
  labs(x = "Run") +
  theme(text = element_text(size = 25))
```
]

.large[
.question[
How else could I compare the observed data to the simulated data?
]
]

---

### Compare new dataset to observed dataset

.large[
$$Metabolic_{ij} = \beta_0 + u_i + \varepsilon_{ij} \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2), \hspace{0.2cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

**Fitted random intercepts model (observed data):**

$\widehat{\beta}_0 = 0.175$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00013$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0042$, $\ \ \widehat{\rho}_{group} = 0.03$

**Fitted random intercepts model (simulated data):**

$\widehat{\beta}_0 = 0.169$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00015$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0049$, $\ \ \widehat{\rho}_{group} = 0.03$

.question[
Do you think there is systematic variation between runs, or do you think the observed differences between runs are due to chance?
]
]

---

### Compare new dataset to observed dataset

.large[
$$Metabolic_{ij} = \beta_0 + u_i + \varepsilon_{ij} \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2), \hspace{0.2cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$

**Fitted random intercepts model (observed data):**

$\widehat{\beta}_0 = 0.175$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00013$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0042$, $\ \ \widehat{\rho}_{group} = 0.03$

**Fitted random intercepts model (simulated data):**

$\widehat{\beta}_0 = 0.169$, $\hspace{0.5cm} \widehat{\sigma}_u^2 = 0.00015$, $\hspace{0.5cm} \widehat{\sigma}_\varepsilon^2 = 0.0049$, $\ \ \widehat{\rho}_{group} = 0.03$

* Estimated variances are similar for observed data and simulated data
* We know there is no random effect in the simulated data
* So maybe observed data isn't unusual if there are no systematic differences between runs
]

---

### Summary (so far)

.large[
.question[
Are there systematic differences between runs (group effects), or are observed differences simply due to chance?
]

* Fit a model with no random effects
* Simulate data from fitted model
* Compare simulated data to observed data
  * If simulated data looks very different, maybe there are systematic differences between runs
  * If simulated data looks similar to observed data, maybe there aren't systematic differences between runs
]

---

### Class activity, Part I

.large[
[https://sta279-s22.github.io/class_activities/ca_lecture_24.html](https://sta279-s22.github.io/class_activities/ca_lecture_24.html)
]

---

### Class activity

.large[
Here are the boxplots and intra-class correlations I got for the music data and the simulated data:

.pull-left[
**Original (observed) data:**

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
music <- read_csv("https://sta279-s22.github.io/labs/music.csv")

music %>%
  ggplot(aes(x = as.factor(id),
             y = na)) +
  geom_boxplot(lwd=1.2) +
  theme_bw() +
  labs(x = "Musician",
       y = "Anxiety")
```

$\widehat{\rho}_{group} = 0.18$
]

.pull-right[
**New (simulated) data:**

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
new_na <- 16.21 +
  rnorm(n=497, mean=0, sd=5.237)

new_data <- data.frame(id = music$id,
                       na = new_na)

new_data %>%
  ggplot(aes(x = as.factor(id),
             y = na)) +
  geom_boxplot(lwd=1.2) +
  theme_bw() +
  labs(x = "Musician",
       y = "Anxiety")
```

$\widehat{\rho}_{group} = 0$

]
]

---

### Class activity

.large[
.pull-left[
**Original (observed) data:**

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
music <- read_csv("https://sta279-s22.github.io/labs/music.csv")

music %>%
  ggplot(aes(x = as.factor(id),
             y = na)) +
  geom_boxplot(lwd=1.2) +
  theme_bw() +
  labs(x = "Musician",
       y = "Anxiety")
```

$\widehat{\rho}_{group} = 0.18$
]

.pull-right[
**New (simulated) data:**

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}

new_data %>%
  ggplot(aes(x = as.factor(id),
             y = na)) +
  geom_boxplot(lwd=1.2) +
  theme_bw() +
  labs(x = "Musician",
       y = "Anxiety")
```

$\widehat{\rho}_{group} = 0$

]

.question[
Do you think there is systematic variation between musicians, or do you think the observed differences between musicians are due to chance?
]
]

---

### Issue

.large[
.pull-left[
**Original (observed) data:**

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
music <- read_csv("https://sta279-s22.github.io/labs/music.csv")

music %>%
  ggplot(aes(x = as.factor(id),
             y = na)) +
  geom_boxplot(lwd=1.2) +
  theme_bw() +
  labs(x = "Musician",
       y = "Anxiety")
```

$\widehat{\rho}_{group} = 0.18$
]

.pull-right[
**New (simulated) data:**

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}

new_data %>%
  ggplot(aes(x = as.factor(id),
             y = na)) +
  geom_boxplot(lwd=1.2) +
  theme_bw() +
  labs(x = "Musician",
       y = "Anxiety")
```

$\widehat{\rho}_{group} = 0$

]

.question[
So far, we have only compared one simulated dataset to the observed data.
]
]

---

### Simulating multiple datasets

.large[
**Plan:** 

* Simulate a dataset
* Compare it to the observed data (calculate $\widehat{\rho}_{group}$ )
* Repeat many times (to get a sense of variability)
]

---

### Simulating multiple datasets

.large[
**Step 1:** Simulate a dataset

```{r include=F}
set.seed(4)
```

```{r}
new_metabolic <- 0.175 +
  rnorm(n=197, mean=0, sd=sqrt(0.0043))

new_data <- data.frame(Run = bugula_early$Run,
                       Metabolic = new_metabolic)
```

.question[
Done!
]
]

---

### Simulating multiple datasets

.large[
**Step 2:** Calculate estimated ICC

```{r}
m_sim <- lmer(Metabolic ~ (1|Run), 
              data = new_data)

summary(m_sim)$varcor
```

$\widehat{\rho}_{group} = \dfrac{0.006^2}{0.006^2 + 0.062^2} = 0.009$

* `summary(m_sim)$varcor` extracts variances of the random effect and residuals

]

---

### Simulating multiple datasets

.large[
**Step 2:** Calculate estimated ICC
]

```{r}
m_sim <- lmer(Metabolic ~ (1|Run), 
              data = new_data)

variance_ests <- as.data.frame(summary(m_sim)$varcor)
icc <- variance_ests[1,4]/(variance_ests[1,4] + variance_ests[2,4])

icc
```

.large[

* `summary(m_sim)$varcor` extracts variances of the random effect and residuals

]

---

### Simulating multiple datasets

.large[
**Step 3:** Repeat many times

* First, we need to create a vector to store the results of our simulations

```{r eval=F}
nsim <- 200  # do 200 repetitions
iccs <- rep(NA, nsim)  # vector to store the results
```

* `nsim` will be our number of simulated datasets
* `iccs` will store the estimated intra-class correlation for each simulated dataset

.question[
What tool do I use in R to repeat something many times?
]
]

---

### Simulating multiple datasets

.large[
**Step 3:** Repeat many times

* Next, we need to iterate with a **for loop**

```{r eval=F}
nsim <- 200  # do 200 repetitions
iccs <- rep(NA, nsim)  # vector to store the results

# repeat simulation multiple times
for(sim in 1:nsim){
  
}
```

* `for(sim in 1:nsim)` means "repeat what follows `nsim` times"

.question[
What goes inside my for loop?
]
]

---

### Simulating multiple datasets


```{r eval=F}
nsim <- 200  # do 200 repetitions
iccs <- rep(NA, nsim)  # vector to store the results

# repeat simulation multiple times
for(sim in 1:nsim){
  new_metabolic <- 0.175 +
  rnorm(n=197, mean=0, sd=sqrt(0.0043))

  new_data <- data.frame(Run = bugula_early$Run, 
                         Metabolic = new_metabolic)
  
  m_sim <- lmer(Metabolic ~ (1|Run), data = new_data)
  
  variance_ests <- as.data.frame(summary(m_sim)$varcor)
  iccs[sim] <- variance_ests[1,4]/(variance_ests[1,4] + 
                                     variance_ests[2,4])
  
}
```


---

### Plotting the results

```{r sim_data, include=F, cache=T}
set.seed(3)

nsim <- 200  # do 200 repetitions
iccs <- rep(NA, nsim)  # vector to store the results

# repeat simulation multiple times
for(sim in 1:nsim){
  new_metabolic <- 0.175 +
  rnorm(n=197, mean=0, sd=sqrt(0.0043))

  new_data <- data.frame(Run = bugula_early$Run, 
                         Metabolic = new_metabolic)
  
  m_sim <- lmer(Metabolic ~ (1|Run), data = new_data)
  
  variance_ests <- as.data.frame(summary(m_sim)$varcor)
  iccs[sim] <- variance_ests[1,4]/(variance_ests[1,4] + 
                                     variance_ests[2,4])
  
}
```

.large[
```{r echo=F, message=F, fig.align='center', fig.width=8, fig.height=3.5}
data.frame(icc = iccs) %>%
  ggplot(aes(x = icc)) +
  geom_histogram(bins = 15) +
  theme_bw() +
  theme(text = element_text(size=25)) +
  labs(x = "Intra-class correlation")
```


.question[
For the observed data, $\widehat{\rho}_{group} = 0.03$. Is this unusual, compared to the simulated data?
]
]

--

.large[
Maybe a little, but not super unusual
]

---

### Plotting the results

.large[
```{r echo=F, message=F, fig.align='center', fig.width=8, fig.height=3.5}
data.frame(icc = iccs) %>%
  ggplot(aes(x = icc)) +
  geom_histogram(bins = 15) +
  theme_bw() +
  theme(text = element_text(size=25)) +
  labs(x = "Intra-class correlation")
```


.question[
How can I summarize how unusual $\widehat{\rho}_{group} = 0.03$ is?
]
]

--

.large[
Calculate the proportion of simulated datasets with $\widehat{\rho}_{group} > 0.03$
]

---

### Summarizing the results

.large[
```{r echo=F, message=F, fig.align='center', fig.width=8, fig.height=3.5}
data.frame(icc = iccs) %>%
  ggplot(aes(x = icc)) +
  geom_histogram(bins = 15) +
  theme_bw() +
  theme(text = element_text(size=25)) +
  labs(x = "Intra-class correlation")
```

```{r}
mean(iccs > 0.03)
```

* The probability of observing $\widehat{\rho}_{group}$ as or more extreme than the correlation from the original data, if there is no systematic variation between runs, is about 0.15
]

---

### Summarizing the results

.large[
```{r}
mean(iccs > 0.03)
```

* The probability of observing $\widehat{\rho}_{group}$ as or more extreme than the correlation from the original data, if there is no systematic variation between runs, is about 0.15

.question[
What do we call this value?
]
]

--

.large[
A p-value! Remember that the p-value is the probability of "our data or more extreme", if the null hypothesis is true
]

---

### Class activity, Part II

.large[
[https://sta279-s22.github.io/class_activities/ca_lecture_24.html](https://sta279-s22.github.io/class_activities/ca_lecture_24.html)
]

---

### Class activity

.large[
```{r eval=F}
nsim <- ...
iccs <- rep(NA, nsim)  # vector to store the results

# repeat simulation multiple times
for(sim in ...){
  new_na <- ...
  new_data <- ...
  
  m_sim <- lmer(na ~ (1|id), data = new_data)
  
  variance_ests <- as.data.frame(summary(m_sim)$varcor)
  iccs[sim] <- variance_ests[1,4]/(variance_ests[1,4] + 
                                     variance_ests[2,4])
}
```

.question[
What should `nsim` be?
]
]

---

### Class activity

.large[
```{r eval=F}
nsim <- 200
iccs <- rep(NA, nsim)  # vector to store the results

# repeat simulation multiple times
for(sim in ...){
  new_na <- ...
  new_data <- ...
  
  m_sim <- lmer(na ~ (1|id), data = new_data)
  
  variance_ests <- as.data.frame(summary(m_sim)$varcor)
  iccs[sim] <- variance_ests[1,4]/(variance_ests[1,4] + 
                                     variance_ests[2,4])
}
```

.question[
What is my range in the for loop? ( `sim in ...` )
]
]

---

### Class activity

.large[
```{r eval=F}
nsim <- 200
iccs <- rep(NA, nsim)  # vector to store the results

# repeat simulation multiple times
for(sim in 1:nsim){
  new_na <- ...
  new_data <- ...
  
  m_sim <- lmer(na ~ (1|id), data = new_data)
  
  variance_ests <- as.data.frame(summary(m_sim)$varcor)
  iccs[sim] <- variance_ests[1,4]/(variance_ests[1,4] + 
                                     variance_ests[2,4])
}
```

.question[
How do I fill in `new_na <- ...` and `new_data <- ...` ?
]
]

---

### Class activity

.large[
```{r sim_music, cache=T, message=F}
nsim <- 200
iccs <- rep(NA, nsim)  # vector to store the results

# repeat simulation multiple times
for(sim in 1:nsim){
  new_na <- 16.21 +
    rnorm(n=497, mean=0, sd=5.237)
  
  new_data <- data.frame(id = music$id,
                         na = new_na)
  
  m_sim <- lmer(na ~ (1|id), data = new_data)
  
  variance_ests <- as.data.frame(summary(m_sim)$varcor)
  iccs[sim] <- variance_ests[1,4]/(variance_ests[1,4] + 
                                     variance_ests[2,4])
}
```
]

---

### Class activity

.large[
```{r echo=F, message=F, fig.align='center', fig.width=8, fig.height=3.5}
data.frame(icc = iccs) %>%
  ggplot(aes(x = icc)) +
  geom_histogram(bins = 15) +
  theme_bw() +
  theme(text = element_text(size=25)) +
  labs(x = "Intra-class correlation")
```

```{r eval=F}
mean(iccs > ...)
```

.question[
How do I fill in the code to calculate a p-value?
]
]

---

### Class activity

.large[
```{r echo=F, message=F, fig.align='center', fig.width=8, fig.height=3.5}
data.frame(icc = iccs) %>%
  ggplot(aes(x = icc)) +
  geom_histogram(bins = 15) +
  theme_bw() +
  theme(text = element_text(size=25)) +
  labs(x = "Intra-class correlation")
```

```{r}
mean(iccs > 0.18)
```

.question[
What do you conclude?
]
]

---

### Class activity

.large[
```{r echo=F, message=F, fig.align='center', fig.width=8, fig.height=3.5}
data.frame(icc = iccs) %>%
  ggplot(aes(x = icc)) +
  geom_histogram(bins = 15) +
  theme_bw() +
  theme(text = element_text(size=25)) +
  labs(x = "Intra-class correlation")
```

```{r}
mean(iccs > 0.18)
```

The observed data, with $\widehat{\rho}_{group} = 0.18$, would be *very* unusual if there were no systematic variability in anxiety between musicians. So we have strong evidence for sysematic variability.
]