---
title: "Lab 2, Question 11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

titanic <- read.csv("https://sta279-s22.github.io/labs/Titanic.csv")

titanic <- titanic %>%
  drop_na()
```

```{r}
titanic_glm <- glm(Survived ~ log(Fare + 1), 
                   family=binomial, data = titanic)
summary(titanic_glm)
```

### Hypotheses

We want to test whether there is a positive relationship between passenger fare and the probability of survival. I chose to use a log transformation on Fare (adding 1 to Fare before transforming because some passengers fares were 0 in the data). A positive relationship between passenger fare and the probability of survival is equivalent to a positive relationship between log(Fare + 1) and the probability of survival. So, our hypotheses are

$H_0: \beta_1 = 0 \hspace{0.5cm} H_A: \beta_1 > 0$

### Test

Because our alternative is one-sided, *we cannot use a likelihood ratio test here*. We have to use a Wald test.

### Test statistic

Our test statistic is $z = \dfrac{\widehat{\beta}_1 - 0}{SE_{\widehat{\beta}_1}} = 0.8089/0.0942 = 8.587$ (where $\widehat{\beta}_1$ and $SE_{\widehat{\beta}_1}$ are taken from the summary R output above).

### p-value

Remember that the p-value R reports in the summary output for the *two-sided alternative* $\beta_1 \neq 0$. Because we have a different alternative hypothesis, this p-value isn't correct.

Here we could either divide the p-value R reports by 2, or calculate it ourself. Let's calculate it ourself:

```{r}
pnorm(8.587, lower.tail=F)
```

So the p-value is approximately 0. Therefore, we have strong evidence that there is a positive relationship between passenger fare and the probability of survival.
