---
title: "Class Activity, March 30"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

Anxiety can affect musicians before performances, and can negatively affect their ability to play and their emotional state. In a [2010 study](https://journals.sagepub.com/doi/10.1177/1948550610370492), researchers examined anxiety in 37 undergraduate music majors. For each musician, data was collected on anxiety levels before different performances (between 2 and 15 performances were measured for each musician). Each row in the data represents one performance, and variables include

* `id`: a unique identifier for the musician
* `na`: negative affect score (a measure of anxiety)
* `large`: whether the musician was performing as part of a large ensemble (large = 1), or as part of a small ensemble or solo (large = 0)

**Question:** Are there systematic differences in anxiety between musicians, or are observed differences due to chance?

To answer this question, we will conduct a simulation, in which we generate new data under the assumption that there are *no* systematic differences in anxiety between musicians.

### Data

The following code imports the data and saves it as a data frame called music:

```{r eval=F}
music <- read_csv("https://sta279-s22.github.io/labs/music.csv")
```

# Part I

Last time, you created a boxplot showing the distribution of anxiety for each musician, and you simulated data from an intercept-only linear model (with no random effect). Here is the code you used last time:

**Boxplot:**

```{r eval=F}
music %>%
  ggplot(aes(x = as.factor(id),
             y = na)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Musician",
       y = "Anxiety")
```

**Fitting models:**

```{r eval=F}
# random intercepts model
m1 <- lmer(na ~ (1|id), data = music)
summary(m1)

# intercept-only linear model
m0 <- lm(na ~ 1, data = music)
summary(m0)
```

**Simulating new data:**

```{r eval=F}
new_na <- 16.21 +
  rnorm(n=497, mean=0, sd=5.237)

new_data <- data.frame(id = music$id,
                       na = new_na)
```

Now that we have our simulated data, we need to compare to the observed data.

1. Create a boxplot showing the distribution of anxiety for each musician in the simulated data.

2. Fit the random intercepts model with the *simulated* data, and report the estimated intra-class correlation.

3. Compare exercises 1 and 2 to your results from the observed data. Does your simulation support the hypothesis that there are systematic differences in anxiety between musicians?


# Part II

Now let's simulate many datasets!

4. Fill in the code below to simulate 200 different datasets from the intercept-only linear model, and store the resulting intra-class correlations (the code may take a minute to run):

```{r eval=F}
nsim <- ...
iccs <- rep(NA, nsim)  # vector to store the results

# repeat simulation multiple times
for(sim in ...){
  new_na <- ...

  new_data <- ...
  
  m_sim <- lmer(na ~ (1|id), data = new_data)
  
  variance_ests <- as.data.frame(summary(m_sim)$varcor)
  iccs[sim] <- variance_ests[1,4]/(variance_ests[1,4] + 
                                     variance_ests[2,4])
  
}
```

5. Make a histogram of the estimated intra-class correlations. To make a quick histogram (without going though `ggplot` ) we can use the `hist` function:

```{r eval=F}
hist(iccs)
```

6. For the original music data, $\widehat{\rho}_{group} = 0.18$. Is this correlation unusual, if the intercept-only linear model is correct? Fill in the code below to calculate a p-value summarizing how unusual the observed data is:

```{r eval=F}
mean(iccs > ...)
```