---
title: "Class Activity, January 31"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

## Setup

In this activity, we will revisit the grad application data. In the slides we looked at GRE as a predictor; here we will use GPA.

We are interested in the logistic regression model for the relationship between GPA and admission to grad school:

$Y_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GPA_i$

1. You are interested in testing whether there is any relationship between GPA and the probability of grad school admission. Carry out both a Wald test and a likelihood ratio test to investigate this question, and report your results. *In practice, we would just do one test, rather than both of them. But since this is a class activity, I want you to try both tests.*

<SCRIPT>
function ShowAndHide(divid) {
    var x = document.getElementById(divid);
    if (x.style.display == 'none') {
        x.style.display = 'block';
    } else {
        x.style.display = 'none';
    }
}
</SCRIPT>

<BUTTON ONCLICK="ShowAndHide('q1_solution')">Solution</BUTTON>

<DIV ID="q1_solution" STYLE="display:none">
**Hypotheses:**

$H_0: \beta_1 = 0 \hspace{0.5cm} H_A: \beta_1 \neq 0$

**Wald test:**

```{r, include=F}
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")
```

```{r}
grad_glm <- glm(admit ~ gpa, data = grad_app, family=binomial)
summary(grad_glm)
```

test statistic: $z = 3.517$. p-value = 0.00043. This p-value is small, so we have strong evidence that there is a relationship between GPA and the probability of admission.

**Likelihood ratio test:**

full model: $\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GPA_i$

reduced model: $\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0$

test statistic: $G = 499.98 - 486.97 = 13.01$

degrees of freedom: 1 (we are testing a single parameter), so $G \sim \chi^2_1$

p-value:

```{r}
pchisq(13.01, df=1, lower.tail=F)
```

</DIV>

2. Instead of the research question in exercise 1, you want to know whether students with a higher GPA have a greater chance of acceptance. Write down your hypotheses in terms of a parameter in the model.

3. Which test (Wald, likelihood ratio, or both) can you use to test the hypotheses in question 2? Carry out the hypothesis test. *Hint: to calculate a p-value for a normal distribution, the `pnorm` function is useful. Use `pnorm(..., lower.tail=F)` for a right-tail alternative, and `pnorm(..., lower.tail=T)` for a left-tail alternative.*

4. Calculate a 95% confidence interval for $\beta_1$ in the logistic regression model.